---
title: "Tidy data: Carleton College Energy Activity"
author: "ECON 122"
date: "Day 8,9"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, eval=FALSE, fig.height = 4, fig.width = 8)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

The data used in this example is from Carleton College. The goal of this activity is to help you better understand what issues arise from working with real world data.

# Building Data
The `AAbuildings` dataset gives the size and year built of buildings on campus that are considered academic or administrative buildings. 

### Question 1
**1.** Run the two read commands below (`read.csv` and `readr::read_csv`). How do they read in the data differently?

**2.** How many buildings on campus are classified as academic or administrative (how many rows are there)?

```{r}
AAbuildings <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/AcadAdminBuildings.csv",stringsAsFactors=TRUE)
glimpse(AAbuildings)
AAbuildings <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/AcadAdminBuildings.csv")
glimpse(AAbuildings)
```

# Energy Use Data
A second dataset gives energy use (kiloWatt hour) every 15 minutes for the 15-16 academic year for all buildings on campus that have an energy meter installed. 

### Question 2
**1.** How have the variable types changed when using `read_csv` instead of `read.csv`?
    - e.g. look at Musser, Timestamp
```{r}
energy1 <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv",stringsAsFactors=TRUE)
glimpse(energy1)
local_edition(1)
energy2 <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv")
glimpse(energy2)
```
**2.** Use the `?read_csv` help file to determine why the `read_csv` command picks character or integer types for some variables. (Hint: look at the `guess_max` argument)

**3.** The Musser Hall readings in `energy2` should not be logical types. Why do the results below suggest that this is an issue?  (You could also look at `problems(energy2)`).  
```{r}
summary(energy1$Musser_Hall)
summary(energy2$Musser_Hall)
```


### Question 3
Read the energy data in again using the `read_csv` command below that specifies column type for `Timestamp` and `dayWeek` and defaults to double types for all other. The order of the factor levels of `dayWeek` are also given so we get days ordered correctly in plots. Note that you will need to wrap a variable name in backticks if it starts with a non-letter character. 

**1.** What are the dimensions of the energy data?

**2.** What do the rows represent? columns represent? Is this "tidy" data (more narrow than wide)?

**3.** How many buildings on campus are in the energy data?

```{r}
energy <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv", 
                   col_type = cols(.default = col_double(), 
                                   Timestamp = col_datetime(format = ""),
                                   dayWeek = col_factor(levels=c("Mon","Tues","Wed","Thurs","Fri","Sat","Sun"))))
glimpse(energy)
summary(energy1$Musser_Hall)  # can check that summaries match
summary(energy$Musser_Hall)
summary(energy$`100_Nevada_Street`)  # use backticks with variable names that start with numbers
```

### Question 4: 

**1.** For the `Center_for_Mathematics_&_Computing`: Use the energy data to plot the average hourly energy usage in October. What trend do you see?

**2.** For the `Center_for_Mathematics_&_Computing`: Show how your plot for #1 above varies by day of the week (`dayWeek`). What trends do you see?

### Question 5
**1.** For the `Center_for_Mathematics_&_Computing`: Use the energy data to plot the average energy usage in Oct. of 2015 by day of the week (`dayWeek`). What trend do you see? 

  - note: if you use `geom_line` you will need to add `group=1` to your `aes` to get a line graph when plotting against the categorical variable `dayWeek`. This is because when we use a `factor` variable `aes` will set the group to each factor. `geom_line` will then try to connect the dots **WITHIN** each group which is just connecting to itself. To fix this, we set `group=1` to tell `ggplot` that all the observations are part of the same `group` and so to connect the line across days of the week. We could also use `group=99`, the main idea is that we set all the observations to the **SAME** group.

**2.** How could you add more buildings to your plot? How easy or hard would this be with the data in the current (wide) format?

# Reshape `energy` data
Create a narrow version of the `energy` data to allow you to *filter* to select one, or more, buildings to plot and analyze. 

### Question 6: 
**1.** What are the dimensions of the narrow version of this data?

### Question 7
**1.** Recreate your graph from Question 5 using the narrow version of this data. What parts of your data wrangling or plot commands do you need to change? 

### Question 8
**1.** Plot a line graph of the average energy usage in Oct. by day of the week (`dayWeek`) for each `building`. What trends do you see? (can you see any?)

  - note: for `geom_line` you will need to add `group=building` to your `aes` to get a line graph for each building when plotting against the categorical variable `dayWeek`

# Joining data
Since we have so many buildings, perhaps limiting our results to academic/admin buildings will help to clear some of the clutter

### Question 9
**1.** How many buildings are in the energy data? 

**2.** How many of the buildings in the energy data are academic/admin buildings?

**3.** How many academic/admin buildings in the building data are not included in the energy data?


### Question 10
Create a dataset by joining the energy and acad/admin building datasets so that you only have acad/admin buildings and full info for both energy and size/year built. 

**1.** Repeat question 8, but this time just for acad/admin buildings. Are any trends or differences in buildings clearer?


### Question 11: use joined data
**1.** Compute the mean energy use by building for Oct. What buildings are the top 5 energy users?

**2.** Repeat the question, but this time compute mean energy use per square foot. Any changes in the top 5?

**3.** Plot the mean energy use per square foot by day of the week for the top 5 buildings found in the previous question and include info about their year built. Compare the daily trends in these buildings. 

