---
title: "Hierarchical Clustering Activity" 
author: "ECON 122"
date: "Day 19"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.width = 9, fig.height = 6)
```


```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr","sparcl","GGally")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Clustering on incoming student characteristics
We will again look at the "classic" college data set of a random sample of colleges and universities. This time we will look at CA **AND** MA schools

```{r}
colleges <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/Colleges.csv")
names(colleges)
colleges2 <- colleges %>% filter(State %in% c("MA","CA"))
table(colleges2$State)
```

We will also focus on the proportion of the incoming class that is in the top 10% or 25% of their HS class.
```{r}
coll_vars <- names(colleges2)[c(3,4,7,8)]
coll_vars
summary(colleges2[,coll_vars])
```
Let's cluster schools by their incoming class characteristics.

#### Question 1
Why should we standardize the variables of interest before using a clustering method?

#### Question 2 
Standardize our four variables of interest and fit a hierarchical clustering model. Plot the dendrogram with label names added.

#### Question 3
Which school seems most similar to University of San Francisco? Compare the SAT and HS class variable values to verify your answer.

#### Question 4
What height would you have to cut this tree to create 3 clusters? Use the `abline(h=)` function to add a cut line to the previous plot.

#### Question 5 
Make 3 clusters and add them as characters to the data frame. Plot a dendogram that identifies each cluster with a different color.

- Hint: Use `cutree`
- Hint: We can add cluster colors to the dendrogram using the `ColorDendrogram` function in the `sparcl` package. 
```{r, eval=FALSE}
ColorDendrogram(coll_hc, y=colleges2$cluster_hc3, labels=colleges2$College, branchlength = 1.6)
```

#### Question 6
What are the characteristics of each cluster? Use either `density curves` or `boxplots` for the `SAT` and `HS` variables. How would you label the clusters in a way the general public would understand? 

#### Question 7
Do you prefer kmeans or hiearchical clustering? What are the pros and cons? 
