---
title: "Iteration"
author: "ECON 122"
date: "Day 11"
output: github_document
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.height = 4, fig.width = 8)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","stringr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Using `sapply()` and `lapply()`


### Question 1
You are considering moving to the following cities and want to gather information about them using the `MonthlyWeather` dataset you are familiar with.

```{r}
weather <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/MonthlyWeather.csv")
cities <- c("Los Angeles","Atlanta","Dallas","Denver","Minneapolis","Pittsburgh")
```

Find the average monthly temperature (across all time periods available) for each city using `dplyr`.

### Question 2 

Find the average monthly temperature (across all time periods available) for each city using `sapply`.

 a. You must first create a function that calculates the `mean` when given a `city` and a `dataset`. Create this function and test it out on one city.

 b. Use `sapply()` to apply your mean function to the vector of cities you are considering. Do you get the same answers as `dplyr`?


### Question 3
While the mean is helpful, you would like to know what the extremes are. Instead of calculating the mean, you would like to create a table that displays the 3 coldest and hottest observations for each city.

 a. Can you perform this task using the simple `dplyr` commands? 

 b. You first need to write a function that outputs a `dataframe` with the 3 coldest and hottest observations given a `city` and `dataset`. Create this function and test it out on one city.

    - **Hint:** `bind_rows()` allows you to combine `dataframes`

 c. Now use `sapply()` to apply your function to our selected cities. Is there an issue? 

 d. What happens if you use `lapply()`? Can you transform your output to create 1 table that has a list of cities and their 3 hottest and coldest temperatures?

 e. Visualize your results by plotting the 3 hottest and coldest monthly temperatures for each city. Which city do you prefer?

