---
title: "Chunk and output options in R Markdown"
author: "ECON 122"
date: "Day 2"
output: github_document
---

**Disclaimer** This .Rmd lacks the usual global chunk options that most/all of my .Rmd files have. This was done to allow you to explore how these options work when compared to the default settings. Since we are exploring R chunk options, to get the most out of this activity you should view the .Rmd file and see the output produced. Just viewing the html is not all that useful! 

## 1. Chunk output options
Here we read in the German credit data and show the breakdown of good/bad loans but don't show the commands:
```{r creditdata, echo=FALSE}
loans <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/CreditData.csv")
table(loans$Good.Loan)
```

Here you show commands that use the `test-assignment` version of the data, but you don't actually evaluate (run) the commands: 
```{r day1creditdata, eval=FALSE}
loans <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/day1CreditData.csv")
table(loans$Good.Loan)
```

The next command we run does evaluate the code. Notice that `loans` is still the full 1000 case dataset, not the 700 case day 1 data. **Why?!**
```{r}
table(loans$Good.Loan)
```
The default chunk options both show code and output.

Notice that the first two chunks are named. We can rerun code in a named chunk just by creating a chunk with the same name but **no** code included. **What happens if we reuse the second chunk, named `day1creditdata`, but this time evaluate the code?**
```{r day1creditdata, eval=TRUE}
```
Notice that we **did not** need to repeat the code in the `day1creditdata` chunk!

Let's return the dataset `loans` to contain full credit dataset. We can reuse the first code chunk but not include any output:
```{r creditdata, include=FALSE}
```

**What chunk options would you set so you only show a boxplot but not the code used to create it?**

## 2. Chunk formatting
The default output look for R chunks highlights code in gray (without `>` prompts), denotes output with two hashtags `##`: 
```{r stats}
summary(loans$Duration.in.month)
tapply(loans$Duration.in.month, loans$Good.Loan, summary)
```

Here we can supress the output hashtags with `comment=NULL`:
```{r stats, comment=NULL}
```
**What happens if you set `comment=";)"`?**

Here we can add `>` in front of code with `prompt=TRUE`:
```{r stats, comment=NULL, prompt=TRUE}
```

Here we can collapse all commands and output in a chunk into one gray command/output field with `collapse=TRUE`:
```{r stats, comment=NULL, prompt=TRUE, collapse=TRUE}
```

## 3. Inline R code
You can use inline R code chunks to place R numerical outout in your text. The full German credit dataset contains `r nrow(loans)` cases. The mean loan duration of these cases is `r mean(loans$Duration.in.month)` months.

You can use the `round` command to get a sensible number of digits (duration is integer valued). The mean loan duration of these cases is `r round(mean(loans$Duration.in.month),1)` months. **Use inline R coding to get the duration standard deviation reported to 1 decimal spot.**

For really big numbers you can use `format` to add commas. The maximum credit amount in the data is  $`r format(max(loans$Credit.amount),big.mark=",")`. 


## 4. Supressing messages
I often load libraries that have annoying messages:
```{r annoying, error=TRUE}
library(dplyr)
```
Check this chunk in your knitted document. You may have an error message in your knitted document (and in your console if you run it there) that tells you to install the package `dplyr`. If so, install the package and reknit your document! Once `dplyr` is installed, you will then get  annoying package message information when loading this library. To supress this message include the option `message=FALSE`.


## 5. Debugging code
If you did not have `dplyr` installed on your computer, then running the `library(dplyr)` command in part 4 would normally cause a "fatal" compilation error that stops your file from knitting. You then get to frantically search your .Rmd file for the location of your error. But by adding the option `error=TRUE` you can get your document to knit and show you the error in the knitted output. Here is an example with `error=TRUE`:
```{r, error=TRUE}
x<- 1:5
(x-1)(x+1)
```
**What is the error in this code?? What happens if you set `error=FALSE`? (Try it and see!)**

One option for debugging a .Rmd file that has R code errors is to set `knitr::opts_chunk$set(error=TRUE)` as a global option at the start of the document. I usually don't start out my Markdown files with `error=TRUE` because it can be too easy to overlook an error in your code this way. 

There are generally two types of errors you get with Markdown docs. If your doc knits after setting `error=TRUE` then you have an R code issue. If the document still does not knit after adding this option, then you have an error in formatting that is not related to your R code. If you have a typo in your chunk options, then you will still get a reference to a chunk line location. Other errors can involve mistakes in your header. **Change your output header type to `guthub_document` to see a non-R chunk type of error message.**


