---
title: "Network Graphs Activity" 
author: "ECON 122"
date: "Day 20"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.width = 9, fig.height = 6)
```


```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr","statnet","circlize","ggnetwork")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Country Migration
The following dataset contains migration counts for decades between 1960 and 2000 between the origin (`origincode`) and destination (`destcode`) countries given in the data. 

```{r}
MigrationFlows <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/MigrationFlows.csv")
MigrationFlows
```

#### Question 1 
Create a subsetted version of this data, called `MigrationFlowsF`, that only contains migration counts of `females` `over 1000` in 2000. How many cases are in this dataset?

#### Question 2
The data frame you created in question 1 can be used as the edge list for a migration network in 2000. Create a `network` object that makes a directional network with edges indicating migration from the origin country to a destination country. How many nodes are in this network? How many edges?

#### Question 3 
Visualizing all countries at once is likely to be overwhelming. Let's analyze the following countries:
```{r}
smallerGroup <- c("USA","CAN","MEX","BRA","CHN","JPN","GBR","PRT")
```
Draw a plot of this network with `ggnetwork` and `ggplot` with vertex labels added. What patterns do you see in the plot? 

- **Note:** Because we have arrows going into and out of the same countries, using `curvature=0.2` in `geom_edges` helps better visualize the flows
- **Note:** Setting a seed `set.seed(x)` can help to keep your network figyre constant each time you run your code

#### Question 4
Is `Y2000`, migration in 2000, an edge or vertex attribute? Add it to your smaller network from question 3 using the `set.xxxx.attribute(network, "name", values)`. 

#### Question 5
Modify your plot from question 3 so that edge width is a function of migration flow size. Use the `size` option in `geom_edges`. What patterns do you notice about migration flows? 

- **Note:** Coloring each `edge` by `country of origin` is helpful to visualize the flows better

#### Question 6 
For a given country, imagine we want to focus on where the inflows of migrants are coming from. Create a new `edge attribute` that calculates for each destination country the ratio of inward migration from relative to all inward migration. Do you learn anything different from this Figure relative to Q5?

