---
title: "Intro to `ggplot2`"
author: "ECON 122"
date: "Day 4"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```


## Trump Tweets
Let's revisit the data collected on Trump tweets.
```{r}
tweets<- read.csv("https://raw.githubusercontent.com/mgelman/data/master/TrumpTweetData.csv")
tweets[226,"charCount"]=76 #modification for density
```

We will focus on the following variables:

Variable | Description
---------| ---------------------------------------
`source` | type of device used to sent the tweet 
`charCount` | how many characters used in the tweet (excluding spaces and weblinks) 
`nWords` | number of "words"" in the tweet (excluding weblinks and retweets)
`nRealWords` | `nwords` minus the number of numbers or "stopwords" (like "a", "the", etc) in the tweet

First look at basic summaries of the variables
```{r}
summary(tweets[,c("source","charCount","nWords","nRealWords")])
```

## Graphs for quantitative variable
The basic graphs for quantitative (numeric) variables are geoms for `histogram`, `density`, `boxplot`, and `points` plot. Let's first just explore the distribution of character count. 

The distribution of character count is
```{r}
library(ggplot2)
w <- ggplot(data=tweets, aes(x=charCount))
w + geom_histogram() + ggtitle("Character Count Distribution") + xlab("Number of characters per tweet")
```

You can change the `binwidth` size to modify bar width:
```{r}
w + geom_histogram(binwidth = 1) + ggtitle("Character Count Distribution")
```

### Question 1: Create a histogram of `nWords`. Note the many tweets with word counts of 0. Word counts of 0 mean the tweet was a retweet or link. Create a histogram of word counts that omits 0 word tweets.

How does character count relate to tweet source? We could compare distributions with a side-by-side boxplot:
```{r}
w <- ggplot(data=tweets, aes(x=source,y=charCount))
w + geom_boxplot() + ggtitle("Character count by source")
```

You can create a horizontal version by calling the `last_plot()` created and flipping axes with `coord_flip()`:
```{r}
last_plot() + coord_flip()
```

You can also compare distributions more thoroughly using density curves for each group. The variable `source` will be used to set the aesthetic `color`:
```{r}
w <- ggplot(data=tweets, aes(x=charCount))
w + geom_density(aes(color=source)) + ggtitle("Character count by source")
```

### Question 2: Why is there a spike in the density? Can you really compare source distributions with this density included?

Recreate the density plot but excluding missing data sources:
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=charCount))
w + geom_density(aes(color=source)) + ggtitle("Character count by source")
```

### Question 3: Do you have any idea why different tweet device sources have different character count distributions?

How do `nWords` and `nRealWords` relate? We could explore this with a scatterplot because both variables are quantitative. We will show how to add both a least squares line and a trend smoother line.
```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_point()
w + geom_smooth()  # adds smoother (loess or gam)
last_plot() + geom_smooth(method="lm", color="red") 
```

### Question 4: Does the relationship look linear? linear over a restricted range of `nWords`? What does adding `scale_x_sqrt()` and `scale_y_sqrt()` to the previous plot do to the plot? Does it make the relationship follow a linear trend better than the original scale?

### Question 5: There are over 1000 tweets in the data. Why don't we see over 1000 glyphs in the plot. In the previous code chunk, what happens when you replace `geom_point()` with `geom_jitter()`?

### Question 6: Are there tweets that are all "real words"? Add `geom_abline(intercept=0,slope=1)` to the original (unjittered) scatterplot to see. 

Does this relationship depend on tweet source? We could explore this visually a couple ways: using a color cue assigned to each glyph or by faceting. 

Here is a plot colored by `source` after omitting the 2 NA sources from the data frame. (The NA `source` group will cause the smoother to fail because there are not enough cases to produce a good smoother trend line.)
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=nWords, y=nRealWords, color=source)) + geom_jitter()
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE)
```

### Question 7: What does `se=FALSE` do (you can try removing it to compare plots)? Why did we have to remove the `color` argument from `geom_smooth`?

### Question 8: Add `size=.4` to the `jitter` geom and `linetype=2` to the `lm` smoother. What changes?

Now we will try faceting by `source`:
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=nWords, y=nRealWords)) + geom_jitter()
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE, color="red") + facet_wrap(~source)
```

### Question 9: Try rerunning the facet chunk without omitting the NA sources. What happens?

### Question 10: Which visual method gives a clearer comparison of the word trends in the three sources?

