---
title: "Data wrangling: multiple data table"
author: "ECON 122"
date: "Day 7"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","nycflights13")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## `nycflights13` package
The `nycflights13` package contains multiple data sets describing `flights` departing from the NYC area in 2013, along with info on `airlines`, `airports`, `planes` and `weather` in 2013. 

#### 1. Take a look at the `flights` `airlines` and `airports` data set help files. 
*hint*: try `?flights` 

- What variable(s) are the keys that connect the `flights` and `airlines` data sets?
- What variable(s) are the keys that connect the `flights` and `airports` data sets?


#### 2. Merging  `airports` and `flights` data. 

- How many *unique* flight `destinations` in `flights` are not given in the `airports` data? (Hint: look at `anti_join` or `semi_join`) Why do you think they are missing?
  - Hint: You may need to use `distinct()`
- How many airports given in `airports` are not `destinations` given in `flights`? Why do you think they are missing? 
- Use `left_join` to add `airports` info to the `flights` data. Why would we want to perform such a join? How many rows are there in this joined data? Then repeat using `inner_join` and `full_join`. Explain the differences in row counts for these three ways of joining.
- Calcuate the median `arr_delay`, average flight `distance`, and number of flights for each `destination`. Plot mean distance vs mean arrival delay, using number of flights to determine point size. Describe the trend you see.
  - *hint*: Use `geom_smooth` to help you see the trend
  - *hint*: Use the `na.rm=TRUE` flag to remove `NA`s
- For fun: Use mean arrival delay by `destination` to map mean delay by location on a map of the US. You can filter `lat` and `long` to restrict attention to the main 48 states. 

#### 3. Use the `weather` data set to add hourly weather information to the `flights` data. 

- The `flights` data contains `hour` and `minute` variables. Do these record the departure, scheduled departure, arrival or scheduled arrival time?
- What variables are the keys to connect `flights` and `weather`?

#### 4. Use the `weather` data set to add the hourly weather information to the `flights` data. 

- Plot the daily average departure delay vs. the daily average precip, using colors to represent origin. Describe the relationship you observe.


