---
title: 'Tidy data: Carleton College Energy Activity - Solution'
author: "ECON 122"
date: "Day 8,9"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.height = 4, fig.width = 8)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr","lubridate")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

The data for this example is from Carleton College. The goal of this activity is to help you better understand what issues arrive from working with real world data.


# Building Data
The `AAbuildings` dataset gives the size and year built of buildings on campus that are considered academic or administrative buildings. 

### Question 1
**1.** Run the two read commands below (`read.csv` and `readr::read_csv`). How do they read in the data differently?

**2.** How many buildings on campus are classified as academic or administrative?

```{r}
AAbuildings <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/AcadAdminBuildings.csv",stringsAsFactors=TRUE)
glimpse(AAbuildings)
AAbuildings <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/AcadAdminBuildings.csv")
glimpse(AAbuildings)
```

#### *Answer:*
**1.** `read.csv` defaults to factor variables for anything with characters while `read_csv` defaults to character variables. 
**2.** There are `r nrow(AAbuildings)` rows in the data set so that is the number of acad/admin buildings. 

# Energy Use Data
A second dataset gives energy use (kiloWatt hour) every 15 minutes for the 15-16 academic year for all buildings on campus that have an energy meter installed. 

### Question 2
**1.** How have the variable types changed when using `read_csv` instead of `read.csv`?
    - e.g. look at Musser, Timestamp
```{r}
energy1 <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv")
glimpse(energy1)
local_edition(1)
energy2 <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv")
glimpse(energy2)
```

**2.** Use the `?read_csv` help file to determine why the `read_csv` command picks character or integer types for some variables. (Hint: look at the `guess_max` argument)

**3.** The Musser Hall readings in `energy2` should not be logical types. Why do the results below suggest that this is an issue?  (You could also look at `problems(energy2)`).  
```{r}
summary(energy1$Musser_Hall)
summary(energy2$Musser_Hall)
```
#### Answer:
**1.** With `read_csv`, `Timestamp` is  a date variable type and `dayWeek` is a character while they were both factors with `read.csv`. There are other numeric variables that are logical valued with `read_csv` and double valued with `read.csv`.

**2.**  `read_csv` guesses at column type by looking at the first 1000 rows (`guess_max` value). Musser Hall has no readings for the first 1000 rows so it incorrectly guesses character. 

**3.** When a numeric was encountered in the Musser Hall data, it is converted to an `NA` value. The `summary` command on the `read_csv` version shows 35129 NA's while the `read.csv` (double) version shows only 22669 NA's. Therefore, `read.csv` correctly imports the column as `numeric` while `read_csv` incorrectly assumes it is logical from the fact that the first 1000 rows are `NA`. The R documentation shows that `logical` is the prefered data type when a column consists of all `NA`s.

### Question 3
Read the energy data in again using the `read_csv` command below that specifies column type for `Timestamp` and `dayWeek` and defaults to double types for all other. The order of the factor levels of `dayWeek` are also given so we get days ordered correctly in plots. Note that you will need to wrap a variable name in backticks if it starts with a non-letter character. 

**1.** What are the dimensions of the energy data?

**2.** What do the rows represent? columns represent? Is this "tidy" data (more narrow than wide)?

**3.** How many buildings on campus are in the energy data?

```{r}
energy <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv", 
                   col_type = cols(.default = col_double(), 
                                   Timestamp = col_datetime(format = ""),
                                   dayWeek = col_factor(levels=c("Mon","Tues","Wed","Thurs","Fri","Sat","Sun"))))
glimpse(energy)
summary(energy1$Musser_Hall)  # can check that summaries match
summary(energy$Musser_Hall)
summary(energy$`100_Nevada_Street`)  # use backticks with variable names that start with numbers
```

#### *Answer:*
**1.** There are `r nrow(energy)` rows and `r ncol(energy)` columns in the data

**2.** The rows are a unique date/hour/minute combination (minutes recorded every 15 minutes) for the 15-16 year. Columns are mainly composed of different academic building energy usage amounts, with some time and date variables. This is very wide data with most columns containing energy usage in different buildings. 

**3.** The first 8 columns are time/date variables, so there are `r ncol(energy)-8` buildings.


### Question 4
**1.** For the `Center_for_Mathematics_&_Computing`: Use the energy data to plot the average hourly energy usage in Oct. What trend do you see?

**2.** For the `Center_for_Mathematics_&_Computing`: Show how your plot for #1 above varies by day of the week (`dayWeek`). What trends do you see?

#### *Answer:*
**1.** The average hourly usage in Oct. will be the average of all hourly readings for all 31 days in Oct. For example, at noon we get 4 readings (15-min increments) so over the entire month the average noon usage is the average of `r 4*31` readings. We get this data by filtering Oct and grouping by hour, then averaging (computing `n()` just to make sure our grouping is correct). Then we can use a line plot to view average usage over the day. In the CMC, we see a peak in usage during classes, then a slight drop at dinner time (~6pm) and another smaller peak around 8pm.
```{r Q4a}
energy_time <- energy %>% 
  filter(month==10) %>% 
  group_by(timeHour) %>% 
  summarize(N=n(),mean_energy=mean(`Center_for_Mathematics_&_Computing`,na.rm=TRUE)) %>%
  print(n=5)
energy_time %>%  ggplot(aes(x=timeHour,y=mean_energy)) + 
  geom_line() + 
  labs(title="Q4 first plot",x="hour of the day",y="average hourly KWH in October")
```

**2.** Same idea as above, but now we need to also group by day of the week. So we are averaging the 4 hourly usage amounts over the 4-5 times a each day of the week occurs in a month. We can see a distinct difference in hourly usage on weekdays vs. weekends. On weekends there is only a small bump in usage during the afternoon and Sunday evening.
```{r Q4b}
energy_daytime <- energy %>% 
  filter(month==10) %>% 
  group_by(dayWeek,timeHour) %>% 
  summarize(N=n(),mean_energy=mean(`Center_for_Mathematics_&_Computing`,na.rm=TRUE)) %>%
  print(n=5)
energy_daytime %>%  ggplot(aes(x=timeHour,y=mean_energy,color=dayWeek)) +
  geom_line() + 
  labs(title="Q4 second plot",x="hour of the day",y="average hourly KWH in October")
```

### Question 5
**1.** For the `Center_for_Mathematics_&_Computing`: Use the energy data to plot the average energy usage in Oct. of 2015 by day of the week (`dayWeek`). 

  - note: if you use `geom_line` you will need to add `group=1` to your `aes` to get a line graph when plotting against the categorical variable `dayWeek`

**2.** How could you add more buildings to your plot? How easy or hard would this be with the data in the current (wide) format?

#### *Answer:*
**1.** Now our analysis unit is day of the week (instead of hour). There are 96 readings per day and there will be 4 to 5 days of the week in a month, so we should see either `r 4*96` or `r 5*96` readings averaged to get our mean usage by day of the week in Oct.  We see that usage peaks mid-week and falls during the weekend.

```{r Q5a}
energy_dayWeek <- energy %>% 
  filter(month==10) %>% 
  group_by(dayWeek) %>% 
  summarize(N=n(),mean_energy=mean(`Center_for_Mathematics_&_Computing`,na.rm=TRUE)) %>%
  print(n=5)  
energy_dayWeek %>%  ggplot(aes(x=dayWeek,y=mean_energy,group=1)) + 
  geom_line() + 
  labs(title="Q5 plot",x="day of the week",y="average daily KWH in October")
```

**2.** In wide form, adding another building is challenging. You could add a couple additional buildings by creating more building specific means during the `summarize` command. You then add these lines to the existing plot by adding more `geom` commands. But to do this for many buildings would require a lot of code, a `for` loop or use of an `apply` command (which we haven't learned yet). The easier way to do this would be to restructure the data.

```{r Q5b}
energy_dayWeek2 <- energy %>% 
  filter(month==10) %>% 
  group_by(dayWeek) %>% 
  summarize(N=n(),mean_energyCMC=mean(`Center_for_Mathematics_&_Computing`,na.rm=TRUE),mean_energyLaird=mean(Laird_Hall,na.rm=TRUE)) %>%
  print(n=5)  
energy_dayWeek2 %>%  ggplot(aes(x=dayWeek,group=1)) + 
  geom_line(aes(y=mean_energyCMC)) + 
    geom_line(aes(y=mean_energyLaird), color="red") +
  labs(title="Q5 plot",x="day of the week",y="average daily KWH in October")
```

