---
title: "Intro to Classifiers - solution" 
author: "ECON 122"
date: "Day 13"
output: 
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.width = 9, fig.height = 4)
```

```{r packageCheck, include=FALSE}
# run the update below in the console if you get an error with str_view
# update.packages(oldPkgs = "stringr", ask=FALSE, repos = "http://cran.us.r-project.org")
mypacks <- c("ggplot2","dplyr","readr","tidyr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Example: Predicting loan defaults
A loan will default if either criteria below is met:

- *Duration is longer than 2 years and credit amount is greater than 10,000 DM.*
- *Duration is 2 years or less and credit amount is less than 2200 DM.*

```{r}
loans <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/CreditData.csv")
loans <- loans %>%
  mutate(c1 = Duration.in.month > 36 & Credit.amount > 10000, 
         c2 = Duration.in.month <= 36 & loans$Credit.amount < 2200, 
         prediction = ifelse(c1 | c2, "BadLoan", "GoodLoan") ) 
(conf.mat <- table(loans$Good.Loan, loans$prediction))
```

#### Question 1: 
What is the confusion matrix if you used 36 months as your duration criteria value? What are the model evaluation stats (accuracy, precision, recall)?

#### *Answer:*

```{r}
(140+346)/1000 # accuracy
140/(140 + 354) # precision
140/(140+160) # recall
loans %>% summarize(
            accuracy = mean(Good.Loan == prediction), 
            precision = sum(Good.Loan == "BadLoan" &  
                              prediction == "BadLoan")/sum(prediction == "BadLoan"),
            recall = sum(Good.Loan == "BadLoan" & 
                     prediction == "BadLoan")/sum(Good.Loan == "BadLoan"))
```


## Random guessing
Suppose you have $n$ observations that you want to predict. We know that $n_0$ are actually failures and $n_1$ are actually successes.

#### Question 2
You flip a fair coin to determine successes (head) and failures (tails) for each case. Write down the confusion matrix that you would expect under this scenario. I.e. what are the expected numbers in each of the four cells?

#### *Answer:*
Of all $n$ cases, half will be predicted to be failures and half success with a fair coin flip predicting outcomes. So
$$
\hat{n}_0 = \hat{n}_1 = \dfrac{n}{2}
$$
Using the same logic, half of the actual failures will be successes and failures:
$$
TN = FP = \dfrac{n_0}{2}
$$
and half of the actual successes will be successes and failures:
$$
FN = TP = \dfrac{n_1}{2}
$$

result | predicted fail | predicted success | total 
---- | ---- | ---- | ----
fail (negative) | $TN=\dfrac{n_0}{2}$ | $FP=\dfrac{n_0}{2}$ | $n_0$ 
success (positive) | $FN= \dfrac{n_1}{2}$ | $TP= \dfrac{n_1}{2}$ | $n_1$ 
total | $\hat{n}_0= \dfrac{n}{2}$ | $\hat{n}_1= \dfrac{n}{2}$ | $n$

#### Question 3
Use your confusion matrix from question 2 to compute accuracy, precision, and recall for this random (fair) guessing prediction method.

#### *Answer:*
With a fair coin flip predicting successes, the accuracy will be 50%. 
$$
accuracy = \dfrac{TN + TP}{n} = \dfrac{0.5n_0 + 0.5n_1}{n} = 0.5 \dfrac{n_0+n_1}{n} = 0.5
$$
Precision will be equal to the rate of successes in the sample since 50% of the successes are correctly predicted and 50% of the overall sample is predicted to be successes:
$$
precision = \dfrac{TP}{\hat{n}_1} = \dfrac{0.5n_1}{0.5n} =  \dfrac{n_1}{n} 
$$
Recall will be 50% since that is the rate at which we correctly predict successes:

$$
Recall = \dfrac{TP}{n_1} = \dfrac{0.5n_1}{n_1} =  0.5 
$$

#### Question 4 
Repeat 2 and 3 but you predict a success with probability $p$ and a failure with probability $1-p$.

#### *Answer:*
Of all $n$ cases, $p$ will be the proportion of successes and $1-p$ the proportion of failures
$$
\hat{n}_1 = pn \ \ \  \ \hat{n}_0 = (1-p)n
$$
Of all $n_0$ failures, $p$ will be the proportion of successes and $1-p$ the proportion of failures
$$
TN = (1-p)n_0 \ \ \ \ FP = pn_0
$$
Of all $n_1$ successes, $p$ will be the proportion of successes and $1-p$ the proportion of failures
$$
FN = (1-p)n_1 \ \ \ \ TP = pn_1
$$

result  | predicted fail | predicted success | total 
--- | --- | --- | ---
fail (negative) | $TN=(1-p)n_0$ | $FP=pn_0$ | $n_0$ 
success (positive) | $FN= (1-p)n_1$ | $TP= pn_1$ | $n_1$ 
total | $\hat{n}_0= (1-p)n$ | $\hat{n}_1= pn$ | $n$


With a probability $p$ predicting successes, the accuracy will be a weighted average of the rates of successes and failures. 
$$
accuracy = \dfrac{TN + TP}{n} = \dfrac{(1-p)n_0 + pn_1}{n} = (1-p)\dfrac{n_0}{n} + p\dfrac{n_1}{n}
$$
Precision will be equal to the rate of successes in the sample since $p$ proportion of the successes are correctly predicted and $p$ proportion of the overall sample is predicted to be successes:
$$
precision = \dfrac{TP}{\hat{n}_1} = \dfrac{pn_1}{pn} =  \dfrac{n_1}{n} 
$$
Recall will be $100p$% since that is the rate at which we correctly predict successes:
$$
Recall = \dfrac{TP}{n_1} = \dfrac{pn_1}{n_1} =  p
$$

#### Question 5
Repeat 2 and 3 but you predict a success with 100% probability! (i.e. all cases are predicted to be a success).

#### *Answer:*
This is question 5 with $p=1$ so

result  | predicted fail | predicted success | total 
--- | --- | --- | ---
fail (negative) | $TN=0$ | $FP=n_0$ | $n_0$ 
success (positive) | $FN= 0$ | $TP= n_1$ | $n_1$ 
total | $\hat{n}_0= 0$ | $\hat{n}_1= n$ | $n$


With a probability $p$ predicting successes, the accuracy will be the rate of successes in the sample:
$$
accuracy = \dfrac{TN + TP}{n} = \dfrac{0 + n_1}{n} = \dfrac{n_1}{n} 
$$
Precision will be equal to the rate of successes in the sample:
$$
precision = \dfrac{TP}{\hat{n}_1} = \dfrac{n_1}{n} 
$$
Recall will be $100$% since that is the rate at which we correctly predict successes:
$$
Recall = \dfrac{TP}{n_1} = \dfrac{n_1}{n_1} =  1
$$

