---
title: "Kmeans Clustering Activity" 
author: "ECON 122"
date: "Day 18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.width = 9, fig.height = 6)
```


```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr","sparcl","GGally")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Clustering using Incoming student characteristics
We will look at a "classic" college data set of a random sample of colleges and universities. For this activity we will focus on CA colleges and universities.

```{r}
colleges <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/Colleges.csv")
names(colleges)
colleges2 <- colleges %>% filter(State=="CA")
```

Instead of focusing just on  `SATM` and `SATV` we will also consider `AppsReceive`. 
```{r}
coll_vars <- names(colleges2)[c(3,4,5)]
coll_vars
summary(colleges2[,coll_vars])
```


First, let's plot the schools against two of the variables `SATM` and `SATV`
```{r}
set.seed(7)
ggplot(colleges2, aes(SATM,SATV)) + 
  geom_point() + 
  geom_text(aes(label=College),position=position_jitter(width=0,height=20),size=2.5) + 
  coord_fixed(xlim = c(400, 800),ylim=c(400,665)) 
```





#### Question 1
Just by looking at the figure above, can you predict roughly what 3 different clusters would look like when using k-means clustering? 

**Ans:**

> A naive answer is that we would expect the clusters to form in groups roughly according to `SATM` and `SATV` groups. However, since we are including `AppsReceive` into our criteria it's not clear how this will affect the clusters. 



#### Question 2 
Use k-means clustering with a `k=3` to cluster the data based on `SATM`, `SATV`, and `AppsReceive`. Add the cluster assignment to your dataset and plot the same figure as above but with colors representing the clusters. Do the clusters match your predictions from Question 1? Why or why not?


**Ans:**

```{r}
km_out <- kmeans(colleges2[,coll_vars], centers=3, nstart=20)
colleges2 <- colleges2 %>% mutate(cluster_assign = as.character(km_out$cluster))

set.seed(7)
ggplot(colleges2, aes(SATM,SATV,color=cluster_assign)) + 
  geom_point() + 
  geom_text(aes(label=College),position=position_jitter(width=0,height=20),size=2.5) + 
  coord_fixed(xlim = c(400, 800),ylim=c(400,665)) 
```

> The clusters don't correspond that well to the spatial position of points in the `SATM` and `SATV` dimension. This is likely due to the fact that `AppsReceive` has a much larger scale and so the distance metric will be dominated by that variable. The clusters will represent distance in this dimension. We can see this by plotting the clusters along the `AppsReceive` dimension on the y-axis


```{r}
set.seed(7)
ggplot(colleges2, aes(SATM,AppsReceive,color=cluster_assign)) + 
  geom_point() + 
  geom_text(aes(label=College),position=position_jitter(width=0,height=20),size=2.5) 
```


#### Question 3
One reason the clustering might have been different from your expectations is because we haven't appropiately scaled the data. Do all the variables have similar scale? Re-do Question 2 using an appropiate scaling. Are the clusters closer to what you were expecting? 

**Ans:**

> The output below shows `AppsReceive` has a larger scale than the other variables. If we scale all the variables, the clusters make more sense when looking at `SATM` vs `SATV`

```{r}
summary(colleges2[,coll_vars])
```

```{r}
km_out_std <- kmeans(scale(colleges2[,coll_vars]), centers=3, nstart=20)
colleges2 <- colleges2 %>% mutate(cluster_assign_std = as.character(km_out_std$cluster))

set.seed(7)
ggplot(colleges2, aes(SATM,SATV,color=cluster_assign_std)) + 
  geom_point() + 
  geom_text(aes(label=College),position=position_jitter(width=0,height=20),size=2.5) + 
  coord_fixed(xlim = c(400, 800),ylim=c(400,665)) 
```



#### Question 4:
Even after standardizing the data, the clusters may seem visually off when plotted on the `SATM` vs `SATV` scale. Why is that? What are the characeristics of the schools that seem to not fit well? 

- Hint: Recreating your plot but using `AppsReceive` on the y-axis may help explain the clusters.

**Ans:**

> Plotting `AppsReceive` against `SATM` shows that the clustering takes into account both `SATM` and `AppsReceive` more equally once we standardize 


```{r}
set.seed(7)
ggplot(colleges2, aes(SATM,AppsReceive,color=cluster_assign_std)) + 
  geom_point() + 
  geom_text(aes(label=College),position=position_jitter(width=0,height=20),size=2.5) 
```

#### Question 5:
Up to now we've been using clusters of 3 to split up the data. Using the standardized data, create a plot that shows how the values of `withinss` (within cluster sum of squared residuals) vary as `k` varies from 1 to 20.

Using this plot, what do you think the optimal level of `k` is? Why? 

> k=3 actually seems pretty optimal. That is where we see the "elbow" in this plot. Higher cluster values will decrease the SS but not by enough to make it worth while (just adds complexity).

```{r}
set.seed(77)
km_twss <- sapply(1:20, function(x) kmeans(scale(colleges2[,coll_vars]), centers=x, nstart=20)$tot.withinss)
km_twss
ggplot(data_frame(k=1:20,km_twss), aes(x=k, y=km_twss)) + geom_point() + geom_line()
```
