---
title: "Iteration"
author: "ECON 122"
date: "Day 11"
output: github_document
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.height = 4, fig.width = 8)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","stringr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Using `sapply()` and `lapply()`


#### Question 1
You are considering moving to the following cities and want to gather information about them using the `MonthlyWeather` dataset you are familiar with.

```{r}
weather <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/MonthlyWeather.csv")
cities <- c("Los Angeles","Atlanta","Dallas","Denver","Minneapolis","Pittsburgh")
```

Find the average monthly temperature (across all time periods available) for each city using `dplyr`.

**ans:**

```{r}
weather %>% filter(city %in% cities) %>% group_by(city) %>% summarise(mean_tmp=mean(avgTmpF))
```


#### Question 2 

Find the average monthly temperature (across all time periods available) for each city using `sapply`.

1. You must first create a function that calculates the `mean` when given a `city` and a `dataset`. Create this function and test it out on one city.

**ans:**

```{r}
meanfun <- function(cityarg,data) {
  data %>% filter(city==cityarg) %>% 
    summarise(mean_tmp=mean(avgTmpF))
}
```

2. Use `sapply()` to apply your mean function to the vector of cities you are considering. Do you get the same answers as `dplyr'?

**ans:**
```{r}
sapply(cities,meanfun,weather)
```

### Question 3
While the mean is helpful, you would like to know what the extremes are. Instead of calculating the mean, you would like to create a table that displays the 3 coldest and hottest observations for each city.

1. Can you perform this task using the simple `dplyr` commands? 

**ans:** Yes, but it is likely going to look quite convoluted. (let me know if you come up with an elegant solution.)

2. You first need to write a function that outputs a `dataframe` with the 3 coldest and hottest observations given a `city` and `dataset`. Create this function and test it out on one city.

  - **Hint:** `bind_rows()` allows you to combine `dataframes`

**ans:**

```{r}
extreme_temps <- function(cityarg,data) {
  bot3 <- data %>% filter(city==cityarg) %>% select(city,avgTmpF) %>% arrange(avgTmpF) %>% head(3)
  top3 <- data %>% filter(city==cityarg) %>% select(city,avgTmpF) %>% arrange(avgTmpF) %>% tail(3)
  bind_rows(bot3,top3)
}

extreme_temps("Atlanta",weather)

```

3. Now use `sapply()` to apply your function to our selected cities. Is there an issue? 

**ans:**
```{r}
str(sapply(cities,extreme_temps,weather))
```
`sapply()` returns a list of 12 but doesn't preserve the `dataframe` structure. We can still recover the data but it is slightly harder to work with

4. What happens if you use `lapply()`? Can you transform your output to create 1 table that has a list of cities and their 3 hottest and coldest temperatures?

```{r}
tlist <- lapply(cities,extreme_temps,weather)
ttable <- tlist %>% bind_rows() 
```

**ans:** The output from `lapply()` is slightly easier to work with because it preserves the dataframe structure. We can simply bind the list together (as seen in the class notes) in order to create one dataframe

5. Visualize your results by plotting the 3 hottest and coldest monthly temperatures for each city. Which city do you prefer?

```{r}
ttable %>% 
  ggplot(aes(x=city,y=avgTmpF)) + geom_point()
```

**ans:** Los Angeles looks very appealing if you ignore the one outlier month that is very hot. 

