---
title: "R Objects Activity - Solution"
author: "ECON 122"
date: "Day 3"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

## Objects used in this handout:
```{r}
x <- c(8,2,1,3)
loans <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/CreditData.csv")
tweets<- read.csv("https://raw.githubusercontent.com/mgelman/data/master/TrumpTweetData.csv")
x.mat2 <- cbind(x,2*x)
x.df <- data.frame(x=x,double.x=x*2)
my.list <- list(myVec=x, myDf=x.df, myString=c("hi","bye"))
```

### Question 1: data types
- What data type is `x`? What data type is `loans$Duration.in.month`? 

```{r}
typeof(x)
typeof(loans$Duration.in.month)
```

- What data type is the vector `c(x, loans$Duration.in.month)`?

**answer:** double is more flexible than integer so the combined vector will be turned into double. 
```{r}
typeof(c(x, loans$Duration.in.month))
```

- What data type is the vector `c(x,"NA")`?

**answer:** The entry `"NA"` is a character string so the vector is coerced into a character vector. To denote a missing value, simply use an unquoted `NA`. 
```{r}
typeof(c(x,"NA"))
typeof(c(x,NA)) # missing value added, still a double
```


### Question 2: subsetting and coercion - may be multiple correct solutions
- How can we get an `x` vector with entries 1 and 2 without using the numbers 1 and 2?
```{r}
x[-c(3,4)]
```
- How can we reverse the order of entries in `x`?
```{r}
x[4:1]
```
- What does `which(x < 5)` equal? 
**answer:** gives index (location) values of entries less than 5
```{r}
x < 5
which(x < 5)
```

- What does `sum(c(TRUE,FALSE,TRUE,FALSE))` equal?
**answer:** number of TRUE's in the vector
```{r}
sum(c(TRUE,FALSE,TRUE,FALSE))
```

- What does `sum(x[c(TRUE,FALSE,TRUE,FALSE)])` equal?
**answer:** first, subset `x` to be just the first and third entries, then sum up these entry values
```{r}
x[c(TRUE,FALSE,TRUE,FALSE)]
sum(x[c(TRUE,FALSE,TRUE,FALSE)])
```

- What does `sum(x < 5)` equal?
**answer:** number of entries less than 5
```{r}
x < 5
sum(x < 5)
```

- What does `sum(x[x < 5])` equal?
**answer:** sums the `x` values that are less than 5
```{r}
x[x < 5]
sum(x[x < 5])
```

- Why does `dim(x.mat2[1:2,1])` return `NULL` while `dim(x.mat2[1:2,1:2])` returns a dimension?
  - **answer:** The first gives the first two rows of the first column of a matrix. Taking only one column from a matrix results in a one-dimensional vector (not a matrix) that no longer has a dimension. The second subset still has more than one row and column, so it is still a matrix and hence has a dimension.
```{r}
class(x.mat2[1:2,1])
dim(x.mat2[1:2,1])
class(x.mat2[1:2,1:2])
dim(x.mat2[1:2,1:2])
```


### Question 3: Data frames 
```{r}
str(tweets$text)
```
- The data set `TrumpTweetData.csv` contains data collected on about 1500 Trump tweets. The variable `text` contains the text of the selected tweets. After reading data in with `read.csv`, R thinks the `text` variable is a factor. Does this make sense? (e.g. would you treat this as a categorical grouping variable in any analysis?)   
  - **answer:** No it doesn't really make sense for it to be a factor.
  
- Use **two methods** to find the `text` of the 180th tweet in the Trump data.
```{r}
tweets$text[180]
tweets[180,"text"]
```

- What is the class of the `attributes` of the data frame `tweets`?
```{r}
class(attributes(tweets))
```


### Question 4: Lists - may have more than one correct answer
- Using `my.list`, show three ways to write one command that gives the 3rd entry of variable `x` in data frame `myDf`
```{r}
my.list
my.list$myDf$x[3]
my.list[["myDf"]]$x[3]
my.list[[2]][3,1]
```

- What class of object does the command `my.list[3]` return?
**answer:** list
```{r}
class(my.list[3])
```

- What class of object does the command `my.list[[3]]` return?
**answer:** (atomic) character vector
```{r}
class(my.list[[3]])
```

- What class of object does the command `unlist(my.list)` return? Why are all the entries `character`s?
**answer:** (atomic) character vector 
```{r}
class(unlist(my.list))
```


### Question 5: Loans revisited
- Give meaning to the following statistical summaries of the loans data from the test-assignment (interpret the numbers given!):
    - `mean(loans$Good.Loan == "BadLoan")` **answer:** proportion of all loans that are "bad loans" (defaults)
    - `mean(loans$Duration.in.month <= 24)` **answer:** proportion of all loans that are less than or equal to 24 months duration
    - `mean(loans$Duration.in.month[loans$Good.Loan == "BadLoan"] <= 24)`  **answer:** proportion of *all loans that defaulted* that are less or equal to 24 months duration.
    
- Explain what the following `ifelse` command produces. 
**answer:** creates a new variable that predicts bad loan for cases over 24 months duration and over 10,000 in credit. 
```{r}
loans$pred.Default1 <- ifelse(loans$Duration.in.month > 24 & loans$Credit.amount > 10000, "predBad", "predGood")
head(loans[, c("Duration.in.month","Credit.amount","pred.Default1")], 10)
```
- Explain what the following `ifelse` command produces.
**answer:** creates a new variable that predicts bad loan for cases over 24 months duration and less than 2,200 10,000 in credit OR for cases less than or equal to 24 months and less than 2,000 in credit. 

```{r}
loans$pred.Default <- ifelse(loans$Duration.in.month <= 24 & loans$Credit.amount < 2200, "predBad", loans$pred.Default1)
head(loans[, c("Duration.in.month","Credit.amount","pred.Default1","pred.Default")], 10)
```
- What data type do the `ifelse` commands above produce (factor or character)? **answer:** character


### Question 6: Functions
```{r}
MeanSD <- function(x,plot=FALSE,...)
{
  mean.x <- mean(x,...)
  sd.x <- sd(x,...)
  if (plot) 
    hist(x)
  return(list(Mean=mean.x,SD=sd.x))
}
```

- Use the `MeanSD` function to get mean, sd and histogram for `loans$Duration.in.month`
```{r}
MeanSD(loans$Duration.in.month, plot=TRUE)
```

- Why does the first command below return NA's while the second returns mean and SD?
**answer:** The commands `mean` and `sd` produce `NA` when there is a missing value in the data entered. The option `na.rm` is used to computes these stats only for non missing values.
```{r}
MeanSD(c(1,2,3,4,NA))
MeanSD(c(1,2,3,4,NA), na.rm=TRUE)
```
- Change the function above to also include the median in it's statistical summary output.
```{r}
MeanSD2 <- function(x,plot=FALSE,...)
{
  mean.x <- mean(x,...)
  sd.x <- sd(x,...)
  med.x <- median(x,...)
  if (plot) 
    hist(x)
  return(list(Mean=mean.x,SD=sd.x, Median=med.x))
}
MeanSD2(loans$Duration.in.month)
```

