---
title: "Intro to `ggplot2`"
author: "ECON 122"
date: "Day 4"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```


## Trump Tweets
Let's revisit the data collected on Trump tweets.
```{r}
tweets<- read.csv("https://raw.githubusercontent.com/mgelman/data/master/TrumpTweetData.csv")
tweets[226,"charCount"]=76 #modification for density
```

We will focus on the following variables:

Variable | Description
---------| ---------------------------------------
`source` | type of device used to sent the tweet 
`charCount` | how many characters used in the tweet (excluding spaces and weblinks) 
`nWords` | number of "words"" in the tweet (excluding weblinks and retweets)
`nRealWords` | `nwords` minus the number of numbers or "stopwords" (like "a", "the", etc) in the tweet

First look at basic summaries of the variables
```{r}
summary(tweets[,c("source","charCount","nWords","nRealWords")])
```

## Graphs for quantitative variable
The basic graphs for quantitative (numeric) variables are geoms for `histogram`, `density`, `boxplot`, and `points` plot. Let's first just explore the distribution of character count. 

The distribution of character count is
```{r}
library(ggplot2)
w <- ggplot(data=tweets, aes(x=charCount))
w + geom_histogram() + ggtitle("Character Count Distribution") + xlab("Number of characters per tweet")
```

You can change the `binwidth` size to modify bar width:
```{r}
w + geom_histogram(binwidth = 1) + ggtitle("Character Count Distribution")
```

### Question 1: Create a histogram of `nWords`. Note the many tweets with word counts of 0. Word counts of 0 mean the tweet was a retweet or link. Create a histogram of word counts that omits 0 word tweets.

**answer:** Histogram of `nWords` 
```{r}
ggplot(data=tweets, aes(x=nWords)) + 
  geom_histogram()
```

Histogram of `nWords` without 0 word tweets

```{r}
ggplot(data=tweets[tweets$nWords>0,], aes(x=nWords)) + 
  geom_histogram()
```


How does character count relate to tweet source? We could compare distributions with a side-by-side boxplot:
```{r}
w <- ggplot(data=tweets, aes(x=source,y=charCount))
w + geom_boxplot() + ggtitle("Character count by source")
```

You can create a horizontal version by calling the `last_plot()` created and flipping axes with `coord_flip()`:
```{r}
last_plot() + coord_flip()
```

You can also compare distributions more thoroughly using density curves for each group. The variable `source` will be used to set the aesthetic `color`:
```{r}
w <- ggplot(data=tweets, aes(x=charCount))
w + geom_density(aes(color=source)) + ggtitle("Character count by source")
```

### Question 2: Why is there a spike in the density? Can you really compare source distributions with this density included?

**answer:** The spike in density comes from the fact that `NA` has very few observations. It's hard to compare the source distributions with `NA` included because the scale has been adjusted to fit `NA` but not the others.

Recreate the density plot but excluding missing data sources:
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=charCount))
w + geom_density(aes(color=source)) + ggtitle("Character count by source")
```

### Question 3: Do you have any idea why different tweet device sources have different character count distributions?

**answer:** This qusetion is based on an [analysis](http://varianceexplained.org/r/trump-tweets/) of Trump tweets that show the different device sources likely reflect different people tweeting from the same account.

How do `nWords` and `nRealWords` relate? We could explore this with a scatterplot because both variables are quantitative. We will show how to add both a least squares line and a trend smoother line.
```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_point()
w + geom_smooth()  # adds smoother (loess or gam)
last_plot() + geom_smooth(method="lm", color="red") 
```

### Question 4: Does the relationship look linear? linear over a restricted range of `nWords`? What does adding `scale_x_sqrt()` and `scale_y_sqrt()` to the previous plot do to the plot? Does it make the relationship follow a linear trend better than the original scale?

**answer:** The relationship looks linear up until `nWords`=20 then starts to flatten out. Converting the x and y axis to a square root scale makes the plot look more linear because it de-emphasizes the larger values where the relationship is not as linear.

```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_point()
w + geom_smooth() + geom_smooth(method="lm", color="red") + scale_x_sqrt() + scale_y_sqrt()
```


### Question 5: There are over 1000 tweets in the data. Why don't we see over 1000 glyphs in the plot. In the previous code chunk, what happens when you replace `geom_point()` with `geom_jitter()`?

**answer:** While there are over 1000 tweets, many of the observations have the same values (for example `nWords`=20) so they will show up as one point. `geom_jitter()` adds a small amount of statistical noise to these values so that we can identify each different point on the plot.

```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_jitter()
w + geom_smooth() + geom_smooth(method="lm", color="red") 
```

### Question 6: Are there tweets that are all "real words"? Add `geom_abline(intercept=0,slope=1)` to the original (unjittered) scatterplot to see. 

**answer:** Yes, there are tweets that are all "real words." `geom_abline(intercept=0,slope=1)` adds a 45 degree line at y=x so that we can identify which tweets have the same value for `nWords` and `nRealWords`. 

```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_point() 
w + geom_smooth() + geom_smooth(method="lm", color="red") + geom_abline(intercept=0,slope=1)
```


Does this relationship depend on tweet source? We could explore this visually a couple ways: using a color cue assigned to each glyph or by faceting. 

Here is a plot colored by `source` after omitting the 2 NA sources from the data frame. (The NA `source` group will cause the smoother to fail because there are not enough cases to produce a good smoother trend line.)
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=nWords, y=nRealWords, color=source)) + geom_jitter()
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE)
```

### Question 7: What does `se=FALSE` do (you can try removing it to compare plots)? Why did we have to remove the `color` argument from `geom_smooth`?

**answer:** `se=FALSE` turns off the standard error bands in the plot. When you are compaing many lines, the standard error bands can make the plot look messy so it's best to remoev them here. We had to remove the `color` argument from `geom_smooth` because the `color=source` argument is not setting the color of the lines. 

### Question 8: Add `size=.4` to the `jitter` geom and `linetype=2` to the `lm` smoother. What changes?

**answer:** These commands adjust the style of the jitter points as well as the `lm` smoother. These simple changes make the plot much easier to analyze.

```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=nWords, y=nRealWords, color=source)) + geom_jitter(size=0.4)
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE,linetype=2)
```


Now we will try faceting by `source`:
```{r}
w <- ggplot(data=tweets[!is.na(tweets$source),], aes(x=nWords, y=nRealWords)) + geom_jitter()
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE, color="red") + facet_wrap(~source)
```

### Question 9: Try rerunning the facet chunk without omitting the NA sources. What happens?

**answer:** If we don't omit `NA` we get an error because the smoothing algorithm doesn't work with only 2 points. 

```{r}
w <- ggplot(data=tweets, aes(x=nWords, y=nRealWords)) + geom_jitter()
w + geom_smooth(se=FALSE) + geom_smooth(method="lm", se=FALSE, color="red") + facet_wrap(~source)
```

### Question 10: Which visual method gives a clearer comparison of the word trends in the three sources?

**answer:** This is subjective, but I prefer the facet approach rather than the color approach for this particular exercise. 


