---
title: "PS3 - solution"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.height = 4, fig.width = 7)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","babynames","dplyr","nycflights13","lubridate","Lahman","mdsr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
  if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

Push your knitted homework assignment (.Rmd, .md docs, and any other required files or folders) to GitHub by the given deadline. 

Package versions
```{r}
mypacks
lapply(mypacks, packageVersion)
```

### Problem 1: explain command (no R needed)

Consider the data set shown in the table below. Each row represents a class and `classType` of `S` denotes a statistics class and `C` denotes a CS class. The variable `m` counts the number of mac users in the class and `w` counts the number of windows users. 

classType | m | w 
------ | -- | --
C | 10 | 4
C | 3 | 1
C | 7 | 3
S | 2 | 7
S | 7 | 10

What data set will be produced by the following commands? Describe the data set in words and show what it looks like using an R Markdown table (like the one used above) to display the new data set. Assume the original data set is named `mydata`.

**1.**
```{r, eval=FALSE}
mydata %>% 
  filter(classType == "C") %>%
  select(w,m)
```

*answer: 1.*
This data set only looks at the number of mac/windows users in CS classes. 

w | m 
-- | --
4 | 10
1 | 3
3 | 7

**2.**
```{r, eval=FALSE}
mydata %>% 
  mutate(ratioW = w/sum(w))
```

*answer: 2.*
The `sum(w)` denominator equals 25, so the ratios are 4/25, 1/25, etc. The value of 25 gives the total number of windows users in all classes so `ratioW` records the proportion of all windows users who are in a given class. 

classType | m | w | ratioW
------ | -- | -- | ----
C | 10 | 4 | 0.16 
C | 3 | 1 | 0.04
C | 7 | 3 | 0.12
S | 2 | 7 | 0.28
S | 7 | 10 | 0.40

**3.** 
```{r, eval=FALSE}
mydata %>% 
  group_by(classType) %>%
  mutate(ratioW = w/sum(w))
```

*answer: 3.*
The `sum(w)` denominator equals the sum of `w` for each `classType` group. The total for `C` classes is 8 windows users in all CS classes listed and for `S` classes is 17 windows users in stats classes.   `ratioW` records the proportion of all windows users in a given class type (CS or stats) who are in a specific class. 

classType | m | w | ratioW
------ | -- | -- | ----
C | 10 | 4 | 0.50 
C | 3 | 1 | 0.125
C | 7 | 3 | 0.375
S | 2 | 7 | 0.412
S | 7 | 10 | 0.588

**4.** 
```{r, eval=FALSE}
mydata %>% 
  group_by(classType) %>%
  summarize(Y = sum(w+m))
```

*answer: 4.*
We are grouping by `classType` then summarizing the class totals `w+m` with the `sum` function. This will produce a small data frame that gives the total number of CS and stats students:

classType | Y
----- | --
C | 28
S | 26


**5.**
```{r, eval=FALSE}
mydata %>% 
  group_by(classType) %>%
  mutate(X = w+m, Y = sum(w+m))
```

*answer: 5.*
We are grouping by `classType` which will only effect the output for any command that summarizes variables like `sum`. The `mutate` command preserves the original data frame and adds the variable `X` that records the total number of mac and windows users in a given class (row). The `Y` variable sums the values of `X` for all entries in each `classType` (grouping variable). These values are repeated (28 CS students and 26 stats students) for all classes in each `classType`. 


classType | m | w | X | Y
------ | -- | -- | -- | --
C | 10 | 4 | 14 | 28
C | 3 | 1 | 4 | 28
C | 7 | 3 | 10 | 28
S | 2 | 7 | 9 | 26
S | 7 | 10 | 17 | 26

### Problem 2: explain output (no R needed)
Consider the data set in given in Problem 1. Explain why the following command (similar to parts 4 and 5) may not produce the intended output as discussed in class.

```{r, eval=FALSE}
mydata %>% 
  group_by(classType) %>%
  summarize(X = w+m, Y = sum(w+m))
```

*answer:*
The result of a summarize command used after a grouping command should be one value per group. The value of `X` in the command above gives a different total for each class (row) in the original data set. These totals can't be reduced down into `classType` values unless another summarizing function (like `sum`) is applied to them. 

### Problem 3 (give command, no R needed)
Consider the data set in given in Problem 1. Provide the `dplyr` commands that will produce the following data frames:

**1.** Adds the proportion of windows users in each class to the original data frame. 

*answer: 1.*
```{r, eval=FALSE}
mydata %>%
  mutate(propW = w/(w+m))
```

**2.** Gives the `classType` and the proportion of windows users amoung all students within each `classType`.  

*answer: 2.*
```{r, eval=FALSE}
mydata %>% 
  group_by(classType) %>%
  summarize(Y = sum(w)/sum(w+m))
```

### Problem 4: babynames 
Use the `babynames` dataset (see day 6 activity) and `dplyr` to answer the following:

**1.** Create a data set `babynames_Kat` that contains data only on female babies named `Katherine`. What is the value of `nrow(babynames_Kat)`? What does this value represent?

*answer: 1.*
```{r nameTrends_a}
babynames_Kat <- babynames %>% 
    filter(name == "Katherine",sex=="F")
nrow(babynames_Kat)
```
There are `r nrow(babynames_Kat)` entries in this data set. This is how many years since 1880 that at least one female was named Katherine. 

**2.** Repeat part **1**, but this time look for all female entries with variations on Katherine: `Katherine`, `Catherine`, `Katie` or `Kate`. 

*answer: 2.*
```{r nameTrends_b}
babynames_Kat <- babynames %>% 
    filter(name %in% c("Katie","Kate","Katherine","Catherine"),sex=="F")
nrow(babynames_Kat)
```
There are `r nrow(babynames_Kat)` entries in this data set. This is the number of years that at least one female was named Katherine plus the number of years that at least one female was named Catherine plus the same quantities for the other two names. 


**3.** Using your data frame from part **2**, create one graph that shows the how the relative proportion of each of these four names varies over time. What trends do you notice? 

*answer: 3.*
```{r nameTrends_c}
ggplot(babynames_Kat, aes(x=year, y=n, fill=name)) + 
  geom_bar(stat='identity',position="fill",width=1) + 
  ggtitle("Name Trends") + 
  ylab("fraction") 
```

The relative popularity of `Catherine` increases until roughly 1970 until `Katie` and `Kate` start to overatke it. The relative popularity of `Katherine` has stayed pretty constant over the years. 


### Problem 5: restaurant violations
Exercise 4.17 (2nd Ed. 4.11). **Note: The problem asks you to focus on `Manhattan`. Use `dplyr` package and both `point` and `smooth` geometries in your graph to see patterns between number of inspections and median score. What pattern do you notice? 

Hint: You'll need to use na.rm=TRUE to compute the median or an `na.omit()` command when wrangling the data. If you use `na.omit()`, make sure you've `select`ed only the key variables needed in your analysis. You don't want to omit a row with an `NA` if it occurs in a variable you aren't using.

#### *answers:*
There is a positive relationship between number of inspections and the median score up until about 2500 inspections. Perhaps this refelcts the fact that zipcodes with many restaurants tend to care more about their overall reputation as a restaraunt cluster vs zipcodes with only a few restaurants. 


```{r violationsPlot}
data(Violations, package="mdsr")
Violations %>% 
  filter(boro == "MANHATTAN") %>%  # omit rows with NA score
  group_by(zipcode) %>%  
  summarize(med_score=median(score, na.rm=TRUE), N=n()) %>% 
  filter(N>=50) %>%   # at least 50 violations/zipcode
  ggplot(aes(x=N,y=med_score)) + 
  geom_point() + 
  geom_smooth()
```

### Problem 6: flights
MDSR Exercise 4.3 (2nd Ed. 4.14)

Hint: You'll need to use some date functions from the `lubridate` package to create a variable for `week`.

#### *answers:*

```{r}
p8 <- flights %>% 
  group_by(tailnum) %>% 
  summarize(N=n()) %>% 
  arrange(desc(N)) %>%
  print(n=5)
```

Flight tail number `r p8[2,1]` has the most flights from NYC airports in 2013 with `r p8[2,2]` flights.

```{r}
weekly_flights <- flights %>% filter(tailnum %in% p8[2,1]) %>%
  mutate(week=week(time_hour)) %>%
  group_by(week) %>%
  summarize (count=sum(n()))

ggplot(weekly_flights, aes(x=week,y=count)) + geom_line() + 
  labs(title=paste0("Number of Trips from NYC for Plane ",p8[2,1]," Each week"))

```



### Problem 7: baseball
MDSR Exercise 4.11 (2nd Ed. 4.13). Use the `Lahman` package. The datasets you will need is  `Teams`.

#### *answers:*

The table below shows the 10 most succesful seasons for The Angels. They won the world series in their second best season 2002.

```{r}
Teams %>% filter(teamIDBR %in% c("CAL","ANA","LAA")) %>%
  select(yearID,W,G,WSWin) %>%
  mutate(win_pct=W/G) %>%
  arrange(desc(win_pct)) %>% 
  head(10)
```

It's unlikely that The Angels won a world series outside their top 10 seasons but the following code checks just in case. It shows that indeed they only won the World Series once. 

```{r}
Teams %>% filter(teamIDBR %in% c("CAL","ANA","LAA")) %>%
  select(WSWin) %>% 
  na.omit() %>%
  mutate(anywin=(WSWin=="Y")) %>% 
  summary(anywin)
```


