---
title: "Strings and regular expressions" 
author: "ECON 122"
date: "Day 10"
output: 
  ioslides_presentation:
    incremental: true
    transition: faster
    widescreen: true
    fig_height: 4
    fig_width: 7
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

```{r packageCheck, include=FALSE}
mypacks <- c("dplyr","stringr","readr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Strings and regular expressions  {.build}

- Today we'll discuss how to work with strings
- Oftentimes we need to search and extract certain patterns from strings    
    - We only want to analyze tweets that mention a specific person or event
    - We want to analyze patterns in telephone numbers from various data scraped from the web
- The bulk of this lecture will focus on using `regular expressions` (regexp) to help us search and extract patterns

## Strings  {.build}

- Create strings with single or double quotes:
```{r}
( s1 <- "ECON 122" )
( s2 <- 'ECON 122 is "fun"' )
writeLines(s2)
```
- `s1` and `s2` are both R strings
- `writeLines` shows the string after parsing "special characters"


## Special characters in R  {.build}

- What if I want to use only `double quotes` in `s2`?
```{r, error=TRUE}
( s2 <- "ECON 122 is "fun"" )   # s2 but with only double quotes
```
- The "escape" backslash `\` is used to escape the special use of certain characters
      - `\"` gives a quote in a string
```{r}
( s3 <- "ECON 122 is \"fun\"" )   # same as s2 string
writeLines(s3)

```

## Special characters in R  {.build}

- So `\` allows us to `escape` the special use of characters but what if we want to use it as a regular old `backslash`?
- What would happen if I run the code  `(s4 <- "Economics\Data Science")`?

```{r, error=TRUE}
(s4 <- "Economics\Data Science")
```

- Need to `escape` the `escape`
```{r}
(s4 <- "Economics\\Data Science")
writeLines(s4)
```

## `stringr` package {.build}

- Another H Wickham package used for extraction of info from strings
- Very useful because all commands start with `str_`
    - In RStudio, you'll see all the commands if you type `str_`

## Combining strings {.build}

- The `stringr` package function `str_c` combines strings with no space:
```{r}
x<- "Apple"
y<- "Pen"
str_c(x,y)
```
- Can specify a separator 
```{r}
(z <- str_c(x,y,sep="-"))
```
- **Note:** Different from `c` which creates a `vector`
```{r}
c(x,y)
```

## Splitting strings {.build}

- `str_split` returns a list of substrings (character vectors)
```{r}
str_split(z, "-")  # separate by space " "
```

## String analysis: Length {.build}

- How many characters in the string:
```{r}
( z <- str_c(x,y,"!", sep=" ") )
str_length(z)
```
- Are spaces and punctuation counted?
- What about special characters?
```{r}
( z <- str_c(x,y,"!", sep="\n") )
str_length(z)
```

## String analysis: Subsetting {.build}

- We can subset strings from the start:
```{r}
(z <- str_c(c("apple","pineapple"), y,sep=" ") )
str_sub(z, start=1, end=4)
```
- or the end:
```{r}
str_sub(z, start=-3, end=-1)
```

## Regular expressions {.build}

- A *regular expression* is a way to describe patterns in strings
    - Terse and powerful, but sometimes hard to read 
    - Not exclusive to R, also integrated into other langauges
- What is the pattern in:
    + time: `09:50:00`
    + words in a sentence: `How now brown cow.`
  
## Finding patterns: practice using `stringr` {.build}

- The helper functions `str_view` and `string_view_all` find regex matches (first and all matches)
```{r}
str_view(z, "p")  # find first "p"
```
```{r}
str_view_all(z, "p")  # find all "p"
```

## Finding patterns using `.` (period) {.build}

- Find a "2"
```{r}
x <- c("the time is 2:10.", "1234")
str_view(x, "2")
```

- Find a "2" and *any* (`.`) character that follows
```{r}
x <- c("the time is 2:10.", "1234")
str_view(x, "2.")
```

- What will `".2."` match?

## Finding patterns using `.` (period) {.build}

- What will `".2."` match?
```{r}
str_view(x, ".2.")
```

- What if we want to find a period `.`?
    - Find a puncuation period: use escape `\\.` regex 
    - Why not `\.`? `.` is not a special character in `R` so it will throw an error
```{r}
x <- c("the time is 2:10.", "the cost is $1.23")
str_view(x, "\\.")
```


## Finding patterns using `[]` to find any occurance {.build}

- Find any numbers between 0 and 9:
```{r}
x <- c("the time is 2:10.", "1234")
str_view_all(x, "[0123456789]")  # same as "\\d"
```

- Numbers between 2 and 4
```{r}
str_view_all(x, "[2-4]")
```

## Finding patterns using `[]` to find any occurance {.build .smaller}

- Explain the following result:
```{r}
str_view_all(x, "[0:9]")
```
```{r}
str_view_all(x, "[\\d:]")
```
```{r}
str_view_all(x, "\\d:")
```
- `[]` is any within

## Finding patterns using `[^]` to avoid occurances {.build}

- Explain the following result:
```{r}
str_view_all(x, "[^\\d:]")
```

- To use `^` for "everything except this", it needs to be used in the brackets `[]`

## Finding patterns using `[^]` to avoid occurances {.build}

- What happens if we forget brackets?
```{r}
str_view_all(x, "^t")
```
```{r}
str_view_all(x, "[^t]")
```

## Finding patterns using anchors (`^` and `$`) {.build}

- Anchors look for matches at the start `^` or end `$`.
- Start or end?
```{r}
str_view(x, "^\\d")
```
```{r}
str_view(x, "\\d$")
```

## Finding patterns using  alternatives {.build}

- Use `|` for `OR` statements
```{r}
str_view(x,"10|12")
```
```{r}
str_view_all(x,".(0|2)")
```

- `()` are used for grouping expressions  
- Similar to mathematical expressions: use them to avoid ambiguity




## Finding patterns using repetition {.build}

- A pattern followed by the number  of matches:
    + `?`: 0 or 1
    + `+`: 1 or more
    + `*`: 0 or more
    + `{n}`: exactly `n` times
    + `{n,m}`: between `n` and `m` times
```{r}
str_view_all(c("color","colour","collar"), "colou?r")
```

## Finding patterns using repetition {.build}

- Find all the numbers in a sentence
```{r}
str_view_all("I love ECON122, I'm sad there are only 15 classes left.","[\\d]+")
```


- Find the last word in a sentence
```{r}
str_view_all("that top cat.", "[a-z]+\\.")
```

## Analyzing strings {.build}

- We can now analyze strings using some common functions
- `str` is a vector of strings and `regex` defines the pattern of interest:
    + `str_detect(str, regex)`: returns logical vector if match
        - `str_subset(str, regex)`: returns `str` entries that match
    + `str_count(str, regex)`: counts the number of detections 
    + `str_replace(str, regex, replstr)`: replaces a pattern with a new one
    + `str_split(str, regex)`: splits a string into a list by `regex` 
    + `str_extract(str, regex)`:  extracts patterns from strings
    + `str_locate(str, regex)`: gives starting and ending position of matches
- extract, replace and locate have `all` versions

## Detecting {.build}

- We can detect the occurrence of substrings within a larger string:
- Q: Does the string "pen" show up in ``r z``?
```{r}
z
str_detect(z, "pen")
```
- Why didn't this work?
- Try again with `str_to_lower`
```{r}
str_detect(str_to_lower(z), "pen")
```

## Splitting {.build}

- split into "words": 
```{r}
str_split("That top cat. It sat on a hat.","\\s") 
```

- split into "sentences":
```{r}
str_split("That top cat. It sat on a hat.","\\.\\s") 
```

## Subset {.build}

- Let's find the "words" that start with t
```{r}
(mywords <- str_split("That arty top cat. It sat on a hat."," "))
(twords <- str_subset(unlist(mywords), "^[Tt][a-z]+"))
```

## Replacing {.build}

- In many cases, non-letter characters can be a nuisance

- how long are the words?
```{r}
(mywords <- unlist(mywords) )
str_length(mywords)
```
- remove non-letter characters: not a letter `"[^[:alpha:]]"`
```{r}
(mywords2 <- str_replace_all(mywords, "[^[:alpha:]]","") )
str_length(mywords2)
```


## Extracting {.build}

- extract all numbers
```{r}
x
str_extract(x, "[\\d]" )
```

## Extracting {.build}

- all numbers  in a string:
```{r}
x
str_extract_all(x, "[\\d]" ) # digits 
str_extract_all(x, "[\\d]+" )  # numbers
```

- what will `"[\\d]{1,2}:[\\d]{2}"` extract?

