---
title: "Iterations with `apply` and `do`" 
author: "ECON 122"
date: "Day 11"
output: 
  ioslides_presentation:
    incremental: true
    transition: faster
    widescreen: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, warning=FALSE, error=FALSE, fig.width = 9)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","stringr","readr","readxl","tidyr","lubridate")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Vectorized language {.build}

- R is a vectorized language
    + avoid loops by applying operations to each element of a vector
```{r}
x <- c(10,20,30,40)
x * -1
```
- The `for` loop version:
```{r}
out <- NULL
for (i in 1:4)
  { out[i] <- x[i]*-1 }
out
```

## Vectorized language {.build}

- Many functions are vectorized
- E.g. `packageCheck` chunk in my slides
    - `install.packages` is a vectorized function
```{r, eval=FALSE}
# what packages are needed?
mypacks <- c("ggplot2","dplyr","stringr","readr","readxl","tidyr") 
# installs each package in the vector mypacks
install.packages(mypacks)  
```
- But `library` is not a vectorized function:
```{r, error=TRUE}
library(mypacks)  # this command fails
```

## Why do we need iteration?

- For the most part, `dplyr` commands `group_by` and `summarize` allow us to calculate various statistics we need
- For more complicated procedures (coming up in second half) we might want to apply a more general `function`. 
    - Ex: Performing a non-standard operation (combining all strings in a group) 
    - Ex: Returing a vector or list of values (10,...,90 %ile) 
    - Ex: Applying a statistical model (least squares,machine learning) 
- Mostly setting us up for the second half of the course

## `apply` family of commands

- R has a family of commands that apply a function to different parts of a vector, matrix or data frame.
    + `apply(matrix, MARGIN, FUN)`: applies the function `FUN` to the matrix `MARGIN` given (1=row, 2=column). Returns an atomic **vector** or **matrix**
        - Ex: Calcualting the mean of all columns (`MARGIN`=2)
    + `lapply(vector, FUN)`: applies the `FUN` to each element in the vector (or list). Returns a list with length equal to that of the vector
        - Ex: Returing a vector or list of values (10,...,90 %ile) 
    + `sapply`: works like `lapply` but returns a vector (so `FUN` can only return one value)
        - Ex: Taking the mean of each group

## `apply`: stats for a data frame {.build .smaller}
-  Read in `EnergyData1516.csv`: Remember columns are buildings
```{r, echo=TRUE}
energy <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv", 
                   col_types = cols(Timestamp = col_datetime(), 
                                    dayWeek = col_factor(levels=c("Mon","Tues","Wed","Thurs","Fri","Sat","Sun")) , 
                                    .default = col_double()))
glimpse(energy)
```

## `apply`: stats for a data frame {.build}

- What if we want median energy usage for all buildings (**columns**) in `EnergyData1516`?
    - KWH values in columns 9:90
    - all columns have the same data type, R will coerce cols `9:90` into a matrix 
    - use **margin 2** to apply the function (`median`) to the columns of the matrix
```{r}
medianKWH <- apply(energy%>%select(9:90), MARGIN=2, median, na.rm=TRUE) 
medianKWH[1:6]
```

## `lapply`: great command for avoiding `for` loops {.build}

- `lapply` is very useful for vectorizing commands that aren't vectorized
    - More general than `apply` because `apply` takes a `vector` or `matrix`
- E.g. R chunk `packageCheck`: 
    + applying the function `library` to each entry in `mypacks`
```{r, eval=FALSE}
# packages to load
mypacks <- c("ggplot2","dplyr","stringr","readr","readxl","tidyr")
# run the library command on all mypack entries: 
lapply(mypacks, library, character.only=TRUE)  
```
- `character.only=TRUE` is a `library` argument

## `lapply`: getting quantiles {.build}

- Let's go to the narrow version of energy

```{r}
energy_narrow <- energy %>% mutate(month = month(month, label=TRUE)) %>%
  gather(key=building, value= energyKWH, 9:90) 
energy_narrow %>% select(building, energyKWH, dayWeek, month) %>% glimpse
```
- Goal: for some grouping of the data, get percentiles 10-90 for each group
```{r, eval=FALSE}
lapply(group_name, myfunction) 
# returns 9 percentiles for each group
```


## `lapply`: getting quantiles {.build}

- We need to crate a function to do this because it is non-standard
- `qfun`: function to create quantiles 
    - input: building string name, data frame
    - output: a data frame with building name, quantiles and quantile identifier
```{r}
qfun <- function(bld, data) {
  # get KWH measures for bld
  kwh <- data %>% filter(building == bld) %>% select(energyKWH)
  # get quantiles .1 thru .9
  q <- quantile(kwh$energyKWH, probs = seq(.1, .9, by = .1), na.rm=TRUE)
  # make data frame
  data_frame(building = bld, 
             q, 
             p = seq(.1,.9,by=.1), 
             qtype = str_c("Perc", 100*seq(.1,.9,by=.1),"%"))
}
```


## `lapply`: getting quantiles {.build}

- Check `qfun` for one building 
```{r}
qfun("Center_for_Mathematics_&_Computing", energy_narrow)
```



## `lapply`: getting quantiles {.build}

- To apply `qfun` to all buildings, need a vector of building names from the `building` variable: 
```{r}
bld_names <- unique(energy_narrow$building)
length(bld_names)
bld_names[1:10]
``` 


## `lapply`: getting quantiles {.build}

- For each entry in the vector `bld_names`, run our quantile function `qfun()`
    + add `qfun` argument `data=energy_narrow` to the `lapply` command
```{r}
qlist <- lapply(bld_names, FUN = qfun, data = energy_narrow)
```
- What is returned?!
```{r}
str(qlist)
```

## `lapply`: getting quantiles {.build}

- `qlist` is a list of 82 data frames, one for each building
```{r}
bld_names[1]
qlist[[1]]
```


## `lapply`: getting quantiles {.build}

```{r}
bld_names[82]
qlist[[82]]
```

## `lapply`: getting quantiles {.build}

- The  `dplyr` command `bind_rows()` coerces a list of similar data frames into one data frame
    - this is why I made `qfun` return a data frame!

```{r}
82*9  # number of rows in qdf
qdf <- qlist %>% bind_rows() %>% print()
```

## `lapply`: getting quantiles {.build}

```{r, fig.height=4}
qdf %>% filter(building %in% c("Sayles-Hill" ,"Language_&_Dining_Center", "Olin_Hall_of_Science"))  %>%
  ggplot(aes(x=q, y=p*100, color=building)) +
  geom_point() + geom_line(aes(group=building)) + labs(y="Percentile (%)",x="KWH") +
  scale_y_continuous(breaks=seq(10,90,by=10))
```

## `do` in `dplyr`{.build}

- the previous `lapply` can't quickly be generalized to other groupings
    - e.g. get quantiles by month? or day of the week?
    - `qfun` needed to `filter` by `building`
- the `do` command in `dplyr` lets you apply a function to the groups of a data frame:
- Idea:
    - write a function `FUN` for a (hypothetical) subsetted version of the data
    - `FUN` must return a data frame or objects in a list 
    - `group_by` to define a subsetted data frame
    - `do(FUN)` runs the `FUN` on the data frames defined by `group_by`

## `do` in `dplyr`: getting quantiles {.build}

- Redo quantile example with modified function (no need to filter groups):
```{r}
qfun2 <- function(data) {
  # get quantiles .1 thru .9
  q <- quantile(data$energyKWH, probs = seq(.1, .9, by = .1), na.rm=TRUE)
  # make data frame
  data_frame(q, 
             p = seq(.1,.9,by=.1), 
             qtype = str_c("Perc", 100*seq(.1,.9,by=.1),"%"))
}
```

## `do` in `dplyr`: getting quantiles {.build}

- group by building 
- then `do(qfun2(data = .))` where `.` will be filled by subsets of the data defined by the groups
```{r}
qdf2 <- energy_narrow %>%
  group_by(building) %>%
  do(qfun2(data = .))
qdf2  # done! same as qdf data!
```

## `do` in `dplyr`: getting quantiles {.build}

- group by building *and* day of the week 
```{r, cache=TRUE}
qdf3 <- energy_narrow %>%
  group_by(building, dayWeek) %>%
  do(qfun2(data = .))
qdf3  
```

## `do` in `dplyr`: getting quantiles  {.build}

```{r, echo=FALSE, fig.height=5.5}
qdf3 %>% filter(building %in% c("Sayles-Hill" ,"Language_&_Dining_Center", "Olin_Hall_of_Science"))  %>%  ggplot(aes(x=q, y=p*100, color=building)) +
  geom_point() + geom_line(aes(group=building)) + labs(y="Percentile (%)",x="KWH") +
  scale_y_continuous(breaks=seq(10,90,by=10)) + facet_wrap(~dayWeek)
```

## Summary

- `apply()` - use if you have a matrix
- `lapply()` - much more general application (ie, can use for `library()`)
- `dplyr::do()` - applying custom functions to the framework of `dplyr` `group_by`
