---
title: "Network statistics"
author: "ECON 122"
date: "Day 21"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, warning=FALSE, error=TRUE, fig.width = 9)
```


```{r packageCheck, include=FALSE}
# if you get an error with ggnetwork, try (re)installing with:
# install.packages("ggnetwork", type="source")
# also install igraph but don't load the library
mypacks <- c("ggplot2","dplyr","readr","tidyr", "statnet","ggnetwork", "intergraph", "gridExtra")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```




## Statistical Analysis of networks
```{r, echo=FALSE, fig.height=5, fig.width=10}
facebook <- readRDS(gzcon(url("https://raw.githubusercontent.com/mgelman/data/master/facebook")))
set.seed(77)
(gface <- ggplot(data=ggnetwork(facebook), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(color="gray50") + 
  geom_nodes(aes(color=sex), size=2) + 
  theme_blank() + coord_fixed() +
  ggtitle("A facebook friends network") )
```


## Statistical Analysis of networks

- Imminent Brewery FB page: post "likers" connected if they like the same post
```{r, echo=FALSE, fig.height=5.5, fig.width=10, cache=TRUE}
facebookImm_Aug <- readRDS(gzcon(url("https://raw.githubusercontent.com/mgelman/data/master/facebookImm_Aug17.RDS"))) 
set.seed(77)
ggplot(data=ggnetwork(facebookImm_Aug), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(color="black", size=.4) + 
  geom_nodes(color="darkred", size=2) + 
  theme_blank() + coord_fixed() +
  ggtitle("Imminent Brewery: Facebook post likers (August 2017)")
```

## Simple example {.build}

```{r, echo=FALSE}
v_data <- data_frame(
  id=c("A","B","C","D","E"),
  sex = c("F","F","M","F","M") )
netlist <- rbind(c("A","B"),
                 c("A","C"),
                 c("B","C"),
                 c("B","D"),
                 c("C","B"),
                 c("E","C"))
mynet <- network(netlist, matrix.type="edgelist")
set.vertex.attribute(mynet, "sex", v_data$sex)
gplot(mynet, displaylabels=TRUE)
```

## Simpsons network

- 15 Simpsons characters with undirected edges defined for characters who interact "a lot"
```{r, fig.height=5, fig.width=10, echo=FALSE}
Simpsons <- readRDS(gzcon(url("https://raw.githubusercontent.com/mgelman/data/master/Simpsons")))
#Simpsons
#gplot(Simpsons, gmode="graph", displaylabels = TRUE, edge.col="lightgray", pad=1)
set.seed(7)
ggplot(data=ggnetwork(Simpsons), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() + 
  geom_nodelabel_repel(aes(label=vertex.names)) + 
  geom_nodes(size=3, color="darkred") +
  theme_blank() + coord_fixed() 
```

## Summary Stats for Networks {.build}

- For most quantitative variables, we can compute summary statistics that inform
    - center (mean, median)
    - spread (sd, IQR, range)
    - skew (quartiles, percentiles) 
- We came compute similar stats for data from nodes or edges
    - e.g. proportion of females in facebook nodes
- Stats can also be used to describe a network's relational data
    - node and edge stats: related to relationships
    - Overall network stats: size, density, clustering, ...
    - unless stated otherwise, we will assume undirected networks


## Node statistics

- **Centrality** stats help assess the *prominence* or *importance* of each node in the network
    - **Degree** - tells us how many edges each node has
    - **Betweenness** - tells us how many important paths cross through that node
    - **Closeness** - tells us how close a node is to other points
    - nodes that stand out with high levels of centrality are the "key players"
-  We can use *summaries* of node-level stats to describe the entire network
    - e.g look at average degree, or the distribution of degree across all nodes

## Node statistics: degree {.build}

- **Degree**: number of edges connected to a vertex
    - directed networks: in, out and total degree counts
- Here we use `degree` from the `sna` package 
    - `sna::degree(network, gmode=)` with gmode `digraph` for directed graphs (default) and `graph` for undirected
    
## Node statistics: degree {.build}

```{r, echo=FALSE, fig.height=3}
set.seed(1)
ggplot(ggnetwork(mynet, arrow.gap=.04), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(arrow=arrow(length=unit(10,"pt"), type="closed")) + geom_nodes() + 
  geom_nodelabel(aes(label=vertex.names)) + coord_fixed() + theme_blank()
```
```{r}
sna::degree(mynet) # total degree
sna::degree(mynet, cmode="indegree") # number edges in
sna::degree(mynet, cmode="outdegree") # number edges out
```
    
## Node statistics: degree {.build}
    
- Add degree as a network node variable    
```{r}
(deg_S <- sna::degree(Simpsons, gmode="graph") )
Simpsons %v% "vertex.names"
set.vertex.attribute(Simpsons, "degree", deg_S)
```
- then vary node size by degree   


## Node statistics: degree {.build}
    
```{r, echo=FALSE}
set.seed(7)
ggplot(data=ggnetwork(Simpsons), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() + 
  geom_nodelabel_repel(aes(label=vertex.names)) + 
  geom_nodes(aes(size=degree), color="darkred") +
  theme_blank() + coord_fixed() 
```

## Node statistics: betweeness {.build}

- **Betweenness**: measures the extent to which a node sits "between" pairs of other nodes. 
    + a high level of betweenness means that a node is in a position to control the flow of "information" in a network

$$
Between(v)= \sum_{s,t\in V\setminus\{v\}}\frac{\sigma_v(s,t)}{\sigma(s,t)}
$$
- $\sigma(s,t)$ # of shortest paths from $s$ to $t$  
- $\sigma_v(s,t)$ # of shortest paths from $s$ to $t$ that pass through $v$


## Node statistics: betweeness {.build}

```{r, include=FALSE}
(btw_S <- betweenness(Simpsons, gmode="graph") )
set.vertex.attribute(Simpsons, "betweeness", btw_S)
Simpsons %v% "vertex.names"
```
- How does degree and betweneess differ?
```{r, fig.height=5, fig.width=10, echo=FALSE}
set.seed(7)
ggplot(data=ggnetwork(Simpsons), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() + 
  geom_nodelabel_repel(aes(label=vertex.names)) + 
  geom_nodes(aes(size=betweeness), color="darkred") +
  theme_blank() + coord_fixed() 
```

## Node statistics: closeness {.build}

- **Closeness**: measures how close a node is to all other nodes in the network
    + a high level of closeness means that only a few steps are needed to get to all other nodes in the network

$$
Close(v)= \frac{1}{avg \ dist}=  \frac{1}{\frac{\sum_{i\neq v}{d(v,i)}}{N-1}}= \frac{N-1}{\sum_{i\neq v}{d(v,i)}} 
$$

- $d(v,i)$: shortest distance between $v$ and $i$
- $N$: the number of nodes


## Node statistics: closeness {.build}

- Gives you a better sense of distance than betweeness (see Smithers)
   
```{r, include=FALSE}
(cls_S <- closeness(Simpsons, gmode="graph") )
set.vertex.attribute(Simpsons, "closeness", cls_S)
Simpsons %v% "vertex.names"
```


```{r, fig.height=5, fig.width=10, echo=FALSE}
set.seed(7)
ggplot(data=ggnetwork(Simpsons), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() + 
  geom_nodelabel_repel(aes(label=vertex.names)) + 
  geom_nodes(aes(size=closeness), color="darkred") +
  theme_blank() + coord_fixed() 
```


## Comparing networks via node statistics {.build}

- You can treat these node stats like any other type of data
- Compare within a network to find unusual cases
    - "central" players
- Compare across networks to compare relational structures
    - may make sense to standardize values to compare extremeness nodes
    
    
## Comparing degree {.build}

- All three networks have at least one high outlier, but different degree scales (different number of nodes)
```{r, echo=FALSE}
degree_fn <- function(net)
{  
  d <- sna::degree(net, gmode="graph")
  data_frame(degree = d)
}
degrees <- lapply(list(facebook, Simpsons, facebookImm_Aug), degree_fn)
degrees[[1]]$name <- "facebook"
degrees[[2]]$name <- "Simpsons"
degrees[[3]]$name <- "Imminent"

degree_df <- bind_rows(degrees)
ggplot(degree_df, aes(x=name, y=degree)) + 
  geom_boxplot()+ 
  ggtitle("Degree distribution")
```


## Comparing degree {.build}

- Highest degree in facebook is over 4 SD's above mean while the largest in Simpsons is just over 2 SD's above.
```{r, echo=FALSE}
degree_fn <- function(net)
{  
  d <- sna::degree(net, gmode="graph")
  data_frame(degree = scale(d)[,1])
}
degrees <- lapply(list(facebook, Simpsons, facebookImm_Aug), degree_fn)
degrees[[1]]$name <- "facebook"
degrees[[2]]$name <- "Simpsons"
degrees[[3]]$name <- "Imminent"

degree_df <- bind_rows(degrees)
ggplot(degree_df, aes(x=name, y=degree)) + 
  geom_boxplot() + 
  ggtitle("Degree distribution (standardized)")
```


## Comparing betweenness {.build}

- Facebooks have a few people with extreme betweenness
```{r, echo=FALSE}
degree_fn <- function(net)
{  
  d <- betweenness(net, gmode="graph") 
  data_frame(betweenness = d)
}
degrees <- lapply(list(facebook, Simpsons, facebookImm_Aug), degree_fn)
degrees[[1]]$name <- "facebook"
degrees[[2]]$name <- "Simpsons"
degrees[[3]]$name <- "Imminent"

degree_df <- bind_rows(degrees)
ggplot(degree_df, aes(x=name, y=betweenness)) + 
  geom_boxplot()+ 
  ggtitle("Betweenness distribution")
```


## Comparing betweenness {.build}


```{r, echo=FALSE, fig.height=6, fig.width=10, cache=TRUE}
set.vertex.attribute(facebookImm_Aug, "between", betweenness(facebookImm_Aug, gmode="graph") )
ggplot(data=ggnetwork(facebookImm_Aug), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(color="black", size=.4) + 
  geom_nodes(aes(size=between), color="darkred") + 
  theme_blank() + coord_fixed() +
  ggtitle("Imminent Brewery: Facebook post likers (August 2017)")
```

## Comparing betweenness {.build}

```{r, echo=FALSE}
ggplot(data_frame(sex = facebook %v% "sex", between = betweenness(facebook, gmode="graph"))) + 
  geom_boxplot(aes(x=sex, y=between)) + 
  scale_y_sqrt() + 
  ggtitle("facebook: betweenness by sex")
```


## Comparing closeness {.build}

- The Facebook network nodes have low closeness, but similar values for all nodes
- Simpsons and Imminent have higher closeness, with two Simpson's outliers 
```{r, echo=FALSE}
degree_fn <- function(net)
{  
  # sna closeness doesn't work for facebook network
  # don't have a good explanation why it doesn't
  d <- igraph::closeness(asIgraph(net), normalized=TRUE)
  data_frame(closeness = d)
}
degrees <- lapply(list(facebook, Simpsons, facebookImm_Aug), degree_fn)
degrees[[1]]$name <- "facebook"
degrees[[2]]$name <- "Simpsons"
degrees[[3]]$name <- "Imminent"

degree_df <- bind_rows(degrees)
ggplot(degree_df, aes(x=name, y=closeness)) + 
  geom_boxplot()+ 
  ggtitle("Closeness distribution")
```

## Comparing closeness {.build}

- Why is closeness so different? 

```{r, echo=FALSE, fig.height=5, fig.width=10, cache=TRUE}
set.vertex.attribute(facebookImm_Aug, "closeness", closeness(facebookImm_Aug, gmode="graph") )
face_cl <- igraph::closeness(asIgraph(facebook), normalized=TRUE)
set.vertex.attribute(facebook, "closeness", face_cl)
g1<- ggplot(data=ggnetwork(facebookImm_Aug), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(color="black", size=.4) + 
  geom_nodes(aes(size=closeness), color="darkred") + 
  theme_blank() + coord_fixed() +
  ggtitle("Imminent Brewery: Facebook post likers (August 2017)")
g2 <- ggplot(data=ggnetwork(facebook), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(color="black", size=.4) + 
  geom_nodes(aes(size=closeness), color="darkred") + 
  theme_blank() + coord_fixed() +
  ggtitle("facebook friends")
grid.arrange(g1,g2, ncol=2)
```


## Network level statistics {.build}

- Some stats describe features of the entire network's relational structure
    - size, number of subcomponents
    - diameter
    - density
    - transitivity (cluster)

## Network level statistics: Size {.build}

- **Size**: how many vertices?
```{r}
network.size(mynet)  
network.size(facebook)
network.size(facebookImm_Aug)
```

## Network level statistics: Subgroups {.build}

- **Number of subgroups**: how many subgroups (connected networks), or components, in undirected networks
```{r, results='hide'}
num_compF <- components(facebook)
num_compImm <- components(facebookImm_Aug)
```
```{r}
num_compF
num_compImm
```
There are 10 subgroups in the facebook friends network and one in the Imminent network

## Network level statistics:  Subgroups {.build}

```{r, results='hide'}
comp_distF <- component.dist(facebook)
```
```{r}
comp_distF$membership[1:10]
comp_distF$csize
comp_distF$cdist[1:10]
```
- The largest subgraph has 47 nodes. 
- There are 3 subgraphs of size one, 1 of size two, etc. 

## Network level statistics:  Diameter {.build}

- **Diameter**: max of the shortest paths across all node pairs
    - max of the largest component
    - measures the compactness of a networks
    - easiest way to get diameter is with `igraph`
```{r}
igraph::diameter(asIgraph(facebook))  # diameter
igraph::diameter(asIgraph(facebookImm_Aug))  # diameter
igraph::diameter(asIgraph(Simpsons))  # diameter
```
Both facebook and Simpons networks only takes at most 4 steps to connect any two nodes, while the Imminent network takes just 3 steps

## Network level statistics:  Diameter  {.build}

- Diameter seems low but central nodes that connect the graph pretty well

```{r, include=FALSE}
facebook_large <- component.largest(facebook, result="graph")
```
```{r, fig.height=5, fig.width=10, echo=FALSE}
set.seed(2)
ggplot(data=ggnetwork(facebook_large), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() + theme_blank() + coord_fixed() + geom_nodes(size=5, color="red")
```


## Network level statistics: Density {.build}

- **Density**: proportion of edges observed in the network to the max number possible
- Let $T$ be the number of ties (edges) in the network with $n$ vertices
$$
den_{directed} = \dfrac{T}{n(n-1)} \ \ \ den_{undirected} = \dfrac{2T}{n(n-1)}
$$
```{r}
network.edgecount(facebook); gden(facebook)
network.edgecount(facebookImm_Aug); gden(facebookImm_Aug)
```
Imminent network has higher density even though network sizes are similar.


## Network level statistics:  Transitivity {.build}

- **Transitivity** is the proportion of closed triangles in the networked to the total number of (open and closed) triangles
$$
Transitivity = \dfrac{3 \times \textrm{# triangles}}{\textrm{# connected triples}}
$$
    - common measure of network **clustering**
    - are your friends also my friends?
```{r}
gtrans(facebook)
gtrans(facebookImm_Aug)
```
Slightly higher clustering in the Imminent network



