---
title: "Network data and graphs"
author: "ECON 122"
date: "Day 20"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, warning=FALSE, error=FALSE, fig.width = 9)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr","readr","tidyr", "statnet","circlize", "ggnetwork")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```




## Statistical Analysis of networks
```{r, echo=FALSE, fig.height=6, fig.width=10}
facebook <- readRDS(gzcon(url("https://raw.githubusercontent.com/mgelman/data/master/facebook")))
set.seed(77)
gplot(facebook, gmode="graph",vertex.col = ifelse((facebook %v% "sex")=="male","lightblue","pink"))
title("A facebook friends network")
legend("bottomright",legend=c("female","male"), col=c("pink","lightblue"), pt.cex=1.5, pch=19)
```

## Networks

- A network consists of vertices (nodes) and edges (ties)
- Vertices: these are usually the "who" or cases in the network
    + people, countries, banks
- Edges: these connect vertices by some defined relationship 
    + people: friendship
    + countries: trade, migration
    + banks: lending, borrowing

## Networks

- Edges may or may not be directional
    + undirected: $A \leftrightarrow B$ 
        - Ann is friends with Bob and Bob is friends with Ann
    + directed:  $A \rightarrow B$ 
        - Ann thinks she is friends with Bob but Bob doesn't consider Ann a friend

## Network data

- Data collected on networks can describe the vertex or edge
- Vertex data
    + variables measured on the vertices like sex, age, country GDP, etc
    + represented in an $n \times p$ data frame ($p$=number of variables)
- Edge data
    + variables that describe the ties between two vertices
    + undirected: $e_{AB}$ measures the amount of time that Ann and Bob spend together in a week
    + directed: value can depend on the direction
        - $e_{AB}$ measures the number of migrants from Algeria to Brazil
        - $e_{BA}$ measures the number of migrants from Brazil to Algeria

## Network data

- Relationship data: how are the nodes connected to each other?
- Two ways to store this information:
    + Adjacency matrix: $n \times n$ matrix with 
        - $(i,j)$ entry = 1 if $i \rightarrow j$, otherwise 0
        - easy to "see", but can be sparse and large for large networks
        - what do we know about the matrix if relationships are undirected? 
    + edge lists: lists the pairs of vertices $(from, to)$ that have edges from the first vertex to the second vertex.
    
## Creating network data in R

- The two common packages to make "network objects" are `network` and `igraph`
- The R package `statnet` is a suite of packages designed to create, graph and analyze network data
    + it includes `network` so that is the package we will use today

## Simple example {.build}

- 5 vertices A-E
- vertex data: sex
```{r}
v_data <- data_frame(
  id=c("A","B","C","D","E"),
  sex = c("F","F","M","F","M") )
v_data
```

## Simple example {.build}

- This adjacency matrix defines our edges:
```{r}
netmat <- rbind(c(0,1,1,0,0),
                c(0,0,1,1,0),
                c(0,1,0,0,0),
                c(0,0,0,0,0),
                c(0,0,1,0,0)) 
rownames(netmat) <- colnames(netmat) <- v_data$id
netmat
```

## Simple example {.build}

- Or we can define the relationship with this edge list of `r sum(netmat)` directional pairs:
```{r}
netlist <- rbind(c("A","B"),
                 c("A","C"),
                 c("B","C"),
                 c("B","D"),
                 c("C","B"),
                 c("E","C"))
netlist
```

## Simple example {.build}

- We take either relationship object as the input to create our network object 
- Here we use the adjacency matrix:
```{r}
mynet <- network(netmat, matrix.type="adjacency")
mynet
```

## Simple example {.build}

- Or via the edge list
```{r}
mynet <- network(netlist, matrix.type="edgelist")
mynet
```

## Simple example {.build}

- Notice that the network contains one vertex variable (attribute): `vertex.name`
    + because the matrix contained row/col names and the list used character names in (from,to) matrix
- You can access vertex data a couple ways
    + using `%v%` operator
```{r}
mynet %v% "vertex.names"
```

- or using `get.vertex.attribute`
```{r}
get.vertex.attribute(mynet, "vertex.names")
```

## Simple example {.build}

- We can add vertex variables using `set.vertex.attributes`:
- Our data frame `v_data` matches our vertex order, so adding `sex` is easy:
```{r}
set.vertex.attribute(mynet, "sex", v_data$sex)
mynet
```

## Simple example {.build}

- We can create a basic plot with `gplot`
```{r}
gplot(mynet, displaylabels=TRUE)
```

## Simple example {.build}

- We can add vertex info: color by `sex`
```{r}
sex_col = ifelse(v_data$sex == "F","yellow","blue")
gplot(mynet, displaylabels=TRUE, vertex.col = sex_col)
```

## Simple example using `ggnetwork` {.build}

- The package `ggnetwork` can create ggplot type network graphs
- The function `ggnetwork` takes in a network objects and produces plotting coordinates:
```{r}
set.seed(1)
(ggdata=ggnetwork(mynet))
```

## Simple example using `ggnetwork` {.build}

- Plot with x/y generated coordinates
```{r, fig.height= 3}
ggplot(ggdata, aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges() +  geom_nodes() + coord_fixed() + ggtitle("Simple example")
```

## Simple example using `ggnetwork` {.build}

- Add directional arrows, node labels and `sex` coloring
```{r, fig.height= 3}
ggplot(ggdata, aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_edges(arrow=arrow(length=unit(10,"pt"), type="closed")) + geom_nodes() + 
  theme_blank() + # turns off arbitrary x/y coordinate choices
  geom_nodelabel(aes(label=vertex.names, color=sex)) +  # sex is a node variable
  coord_fixed() + ggtitle("Simple example")
```

## Facebook example {.build}

- The `network` object was saved as a `RDS` object (single R object) 
- Can read it in with `readRDS`: (note it is undirected)
```{r, echo=TRUE, fig.height=6, fig.width=10}
facebook <- readRDS(gzcon(url("https://raw.githubusercontent.com/mgelman/data/master/facebook")))
facebook
```

## Facebook example {.build}

```{r, fig.height=4}
ggplot(data=ggnetwork(facebook), aes(x = x, y = y, xend = xend, yend = yend)) + 
  geom_nodes(aes(color=sex), size=2) + 
  geom_edges(color="gray50") + 
  theme_blank() + coord_fixed() +
  ggtitle("A facebook friends network")
```

## Graphing Networks {.build}

- Graphs of the same network may have different vertex positions depending on random seed
    + the graph is **relational** and  *usually* not tied to a coordinate system!
    + show structure, important vertices ("actors"), subgroups, etc
    + can also show node, edge info via color, size, shape, etc
    + usually, edge length is arbitrary
- There are many ways to display networks
- `gplot` and `ggnetwork` defaults are both `fruchtermanreingold`
    + example of a **force-directed** graphing algorithm
    + pairs of connected nodes are attracted
    + pairs of unconnected nodes are repulsed
    + built by randomly arranging nodes
    
## Graphing Networks (simple example) {.build}

```{r}
par(mfrow=c(1,2))
gplot(mynet, displaylabels=TRUE, vertex.col = sex_col, mode="random", main="random")
gplot(mynet, displaylabels=TRUE, vertex.col = sex_col, main="force-directed")
```

## Graphing Networks (Facebook) {.build}

- see `?gplot.layout` for layout options
```{r}
sex_colface = ifelse((facebook %v% "sex")=="male","lightblue","pink")
par(mfrow=c(1,2))
gplot(facebook, gmode="graph", vertex.col = sex_colface, mode="random", main="random")
gplot(facebook, gmode="graph", vertex.col = sex_colface,  main="force-directed")
```

