---
title: "R: nuts and bolts" 
author: "ECON 122"
date: "Day 3"
output: 
  ioslides_presentation:
    incremental: false
    widescreen: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

## Where do things live? | **Working Directory** {.build}

- Directory where R first looks for files. 
    + If you run `read.csv("mydata.csv")` then `mydata.csv` should be in your working directory
- Check current location (your location will be different):
```{r}
getwd()
```
- **R Markdown:** When you `knit` a .Rmd file, the working directory for the compilation of the document is always the **directory where the .Rmd file is located!** 
    + This may be different from the location of your current Rstudio session.

## Where do things live? | **Workspace** {.build}

- This is the "environment" where R objects live. 
    + See the **Environment** tab in Rstudio
- Check the contents of my (or your) environment
```{r}
ls()
```
(If it says `character(0)` then your workspace is empty!)

-  **R Markdown:** When you `knit` a .Rmd file, a completely **new and separate working space** is created. 
    + **why do you think this is the case?**
    + R chunk may throw an error if you reference an object loaded in your workspace but not via R chunk
        
    
## Where do things live? | **Rstudio Projects** {.build}

- Projects set your working directory to the folder it lives in
    + easy way to set your working directory
    + can save and reload your workspace (environment) 
- Projects can be connected to GitHub using Rstudio's GUI
- **Highly recommend:** create a project for this class in the folder you are using to store class work.
    + see [Hadley's page on projects](http://r4ds.had.co.nz/workflow-projects.html)

## Objects in R {.build}

- Anything created or imported into R is called an **object**
    + vectors, data frames, matrices, lists, functions, lm, ...
- We usually store objects in the workspace using the assignment **operator** `<-`
```{r}
x <- c(8,2,1,3)
ls()
```
- The `=` operator also does assignment, but it is mainly used for argument specification inside a function.
```{r}
y <- rnorm(3, mean=10, sd=2)
ls()
```
- Please don't use `=` for variable assignment!!!

## Data types | Determines computer storage of info {.build}

- Important data types
    - Logical: TRUE and FALSE are the only values
    - Numeric class: Integer and double
        - default is double
    - Character: string ("") of text 

```{r}
typeof(x)  # type of storage mode 
typeof("abc")
x == 1
typeof(x == 1)
```

## Vectors | Shape of an object {.build}

- R uses two types of vectors to store info
    - **atomic vectors**: all entries have the same data type
    - **lists**: entries can contain other objects that can differ in data type
- All vectors have a length

```{r}
x
length(x)
x.list<- list(x,1,"a")
length(x.list)
```

## Atomic Vectors: Matrices {.build}

- You can add **attributes**, such as **dimension**, to vectors
- A **matrix** is a 2-dimensional vector containing entries of the same type
```{r}
x
x.mat <- matrix(x, nrow=2, byrow=TRUE)
x.mat
class(x.mat)
str(x.mat)
```
- Note: `str` is a handy way of dispalying the structure of objects

## Atomic Vectors: Matrices {.build}

- or you can bind vectors of the same length to create columns or rows:
```{r}
x.mat2 <- cbind(x,2*x)
x.mat2
```


## Lists: Data frames {.build}

- A **data frame** is a list of atomic vectors of the same length, but not necessarily the same data type
- the `loans` data frame has columns that are `integer` and `factor` types
```{r}
loans <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/CreditData.csv",stringsAsFactors=TRUE)
class(loans)
typeof(loans)
str(loans)
```

## Lists: Data frames {.build}

- data frame attributes include column `names` and `row.names`
- you can create a data frame with the `data.frame` command
```{r}
x.df <- data.frame(x=x,double.x=x*2)
x.df
attributes(x.df)  # note: attributes returns a list!
```

## Data types: factors {.build}

- Factors are a data type stored as integers
    - The attribute `levels` is a character vector of possible values
    - Values are stored as the integers (1=first `level`, 2=second `level`, etc)
```{r}
class(loans$Good.Loan)
typeof(loans$Good.Loan)
levels(loans$Good.Loan)
str(loans$Good.Loan)
head(loans$Good.Loan)
```

## Object Oriented Programming {.build}

- In R, commands care about object class and type
- Ex: For a factor, what type of `summary` would be helpful?
```{r}
summary(loans$Good.Loan)
```
- What about for a numeric? 
```{r}
summary(loans$Duration.in.month)
```

- In your **Console** window, type `?summary` then hit **tab**. 
    - see `summary.default`, `summary.factor`, ...


## Coercion {.build}

- Entries in atomic vectors must be the same data type
- R will default to the most complex data type if more than one type is given
```{r}
y <- c(1,2,"a")
y
typeof(y)
```

## Coercion {.build}

- Logical values coerced into 0 for `FALSE` and 1 for `TRUE`
```{r}
z <- c(TRUE, FALSE, TRUE, 7)
z   # TRUE = 1, FALSE = 0
typeof(z)
```
- Logical vectors are also coerced into numeric when applying math functions
```{r}
x
x >= 5  # which entries >= 5?
sum(x >= 5)  # how many >=5 ?
mean(x >= 5) # proportion of entries >=5
```


## Subsetting: Atomic Vector  {.build} 

- subset with `[]` by referencing index value (from 1 to vector length):
```{r}
x
x[c(4,2)]  # get 4th and 2nd entries
```
- subset by omitting entries
```{r}
x[-c(4,2)]  # omit 4th and 2nd entries
```
- subset with a logical vector
```{r}
x[c(TRUE,FALSE,TRUE,FALSE)]  # get 1st and 3rd entries
```

## Subsetting: Matrices {.build}

- access entries using subsetting `[row,column]`
```{r}
x.mat2
x.mat2[,1] # first column
x.mat2[1:2,1] # first 2 rows of first column
```

- R doesn't always preserve class:  
```{r}
class(x.mat2[1,])  # one row (or col) is no longer a matrix (1D)
```

## Subsetting: Data frames {.build}

- you can access entries like a matrix:
```{r}
x.df
x.df[,1]  # first column, all rows
class(x.df[,1])  # first column is no longer a data frame
```

- One column of a data frame is no longer a data frame 

## Subsetting: Data frames {.build}

- but *one row* with 2 or more columns (variables) is still a data frame
```{r}
x.df[1, 1:2]  # 1 row, 2 columns
class(x.df[1, 1:2])
```
- Remember that data frames are built from **lists** of variables (vectors) so they are different objects than matrices

## Subsetting: Data frames {.build}

- or access columns with `$`
```{r}
x.df$x  # get variable x column
```
- you can also use column names to subset:
```{r}
loans[1:2,c("Good.Loan","Credit.amount")] # get 2 rows of Good.Loan and Credit.amount
```


## Subsetting: Lists {.build}

- Recall: a **list** is a vector with entries that can be different object types
```{r}
my.list <- list(myVec=x, myDf=x.df, myString=c("hi","bye"))
my.list
```

## Subsetting: Lists {.build}

- one `[]` operator gives you the object at the given location but preserves the list type
- `my.list[2]` return a **list** of length one with entry `myDf` 
```{r}
my.list[2]
str(my.list[2])
```

## Subsetting: Lists {.build}

- the double  `[[]]` operator  gives you the object stored at that location
    - can enter location number or entry name
- `my.list[[2]]` or `my.list[["myDf"]]` return the **data frame** `myDf` 
```{r}
my.list[[2]]
str(my.list[[2]])
my.list[["myDf"]]
```


## Subsetting: Lists {.build}

- Like a data frame, can use the `$` to access objects stored in the list
    - equivalent to using `[[]]` 
- `my.list$myDf` return the **data frame** `myDf` 

```{r}
my.list$myDf
class(my.list$myDf)
```

## Functions {.build}

- A function takes in objects and arguments and produces a new object
- Function creation is not something we will emphasize in this Data Science course, but you should be able to create simple functions

```
my.function <- function(objects,arguments)
{
  code that does the work
  
  return(list(objects that you are returning))
}
```

## Functions {.build}

- This function returns the mean and SD of a vector and (possibly) creates a histogram
    + input: variable `x`, plotting argument (default is `FALSE`)
    + ...: optional arguments that might be used in `mean` and `sd`
    + output: list of mean and sd
```{r}
MeanSD <- function(x,plot=FALSE,...)
{
  mean.x <- mean(x,...)
  sd.x <- sd(x,...)
  if (plot) 
    hist(x)
  return(list(Mean=mean.x,SD=sd.x))
}
```

## Functions {.build}

- A simple example:
```{r}
MeanSD(loans$Credit.amount)
```
- Function output is a list:
```{r}
str(MeanSD(loans$Credit.amount))
```

## Functions 
<div class="columns-2">
Add plotting argument:
```{r, fig.height=4,fig.width=4}
MeanSD(loans$Credit.amount, plot=TRUE)
```
</div>

