---
title: "`ggplot` graphics" 
author: "ECON 122"
date: "Day 4"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, warning=FALSE, error=TRUE)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Exploratory data analysis (EDA)

- Our ultimate goal is a statistical hypothesis test or robust statistical pattern
  - However, oftentimes, real world data is messy
- Ex: you come up with a thesis topic and write down all the theory and statistical analysis you want to perform
    - You find a relevant dataset, but should you trust it? 
- With large datasets, unrealistic to go through and check all the data
- Plotting a few key relationships can help us determine how usable our data is
- ggplot is your best friend when it comes to EDA! 

## `ggplot2`: a grammar for graphics {.build}

- created by Hadley Wickham
- purposefully written to follow good graphing taxonomy
- allows you to create a graph in small, *readable* code chunks
- install the `ggplot2` package on your computer (if needed)
```{r, message = FALSE}
library(ggplot2)
packageVersion("ggplot2")
```


## `ggplot2` grammar {.build}

- If you are really interested in the theory of Data Visualization, see Ch. 2
    - Basic idea is that there is a hierarchy of how we notice graphical features

- ggplot attempts to follow that hierarchy 
```
ggplot(data) +    # data
  <geometry_funs>(aes(<variables>)) + # aesthetic variable mapping
  <label_funs> +  # add context
  <facet_funs> +  # add facets (optional)
  <coordinate_funs> +  # play with coords (optional)
  <scale_funs> + # play with scales (optional)
  <theme_funs> # play with axes, colors, etc (optional)
```
- can **layer** geometry functions


## `Survey` data {.build}

- the dataset `Survey` provides student data from a survey:
```{r}
survey <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/Survey.csv")
```
- Look first at GPA (Q15) and possible sci or non-sci major (Q16) 



## Data {.build}

- The first layer sets the graphical environment
```{r, fig.height=3.5}
ggplot(data=survey) 
```


## Aesthetics  {.build}

- Aesthetics describes the mapping of variables to your graph
    - for numeric, default scale is Cartesian 
```{r, fig.height=3.5}
ggplot(data=survey, aes(x=Question.15)) 
```

## Geometry  {.build}

- The geometry determines what form the plot has
- What type of plot would be useful here?
    + Here a `histogram` makes sense:
```{r, fig.height=2.5}
ggplot(data=survey, aes(x=Question.15)) + geom_histogram()
```

- also `ggplot(data=survey) + geom_histogram(aes(x=Question.15))`

## EDA for data clean up  {.build}

- Obviously there were some typos in the GPA data. 
- Subset the data so only cases with a reasonable GPA are included.
```{r, fig.height=3.5}
survey <- survey[survey$Question.15 <=4.0 & survey$Question.15 >0,]
ggplot(data=survey, aes(x=Question.15)) + geom_histogram()
```

## Faceting  {.build}

- Can add a third variable, science or non-sci major (Q16), with a facet:
```{r, fig.height=3.5}
ggplot(data=survey, aes(x=Question.15)) + geom_histogram() + facet_wrap(~Question.16)
```

- Note: need `~` before the variable for `facet_wrap()`

## Another option  {.build}

- Could also use side-by-side boxplots:
```{r, fig.height=3.5}
ggplot(data=survey, aes(y=Question.15, x=Question.16)) + geom_boxplot() 
```

- This is where ggplot shines. Very easy to change the plot 

## `ggplot` vs `base r` and other languages

- Typically, plots are based on the plot type
  - for example, `base r` has `hist()` to plot histograms
- As a data scientist, makes sense to think about data and mapping instead of plot
- `ggplot` is a wrapper for `base r` plotting functions 
  - allows data scientists to focus on how to display data rather than a specific plot
  - can initialize the data and then change the plot type easily
  - if you start off with `hist()` and want to change to `boxplot()` all the inputs are different
      - in `ggplot` it's as easy as swapping out `geom_histogram` with `geom_boxplot`

## `ggplot` and `NA`'s  {.build}

- Need to omit `NA` rows from data if you want them omitted from your graphics
```{r, fig.height=3.5}
survey <- survey[!is.na(survey$Question.16),]
ggplot(data=survey, aes(y=Question.15, x=Question.16)) + geom_boxplot() 
```

## Adjusting coordinates  {.build}

- We can flip the x/y coordinates (`boxplot` always wants `x` to be categorical and `y` numeric)
```{r, fig.height=3.5}
ggplot(data=survey, aes(y=Question.15, x=Question.16)) + geom_boxplot() + coord_flip()
```

## Context! Add labels {.build}

- Use `labs` to add labels, `theme` to change text size
```{r, fig.height=3.5}
ggplot(data=survey, aes(y=Question.15, x=Question.16)) + geom_boxplot() + coord_flip() + 
  labs(title="GPA by major", x="Major Area", y="GPA") + 
  theme(text=element_text(size=18))
```

## How to proceed {.build}

- Get to know basic command structure (`ggplot + geom`)
    - Use cheat sheet to see `aes` options for each `geom`
    - `help`...`cheat sheets`
- Then add context layers: labels, font sizes, etc
- `?theme`: non-data features (fonts, legends, axes)
-  `?scale_`: scale (x,y,fill, shape) data features


## Scatterplot {.build}

- The previous example looks at the properties of **one** variable 
- Now, let's look at the relationship between **two** variables
    -  number of facebook friends (Q12) and GPA (Q15) 
```{r, fig.height=3.5}
ggplot(survey, aes(x=Question.12,y=Question.15)) + geom_point() 
```

## Scatterplot {.build}

Change symbol size and shape
```{r, fig.height=3.5}
ggplot(survey, aes(x=Question.12,y=Question.15)) + geom_point(size=3,shape=2) 
```

## Scatterplot {.build}

Change symbol shape by `major` (Q16): requires `aes` argument!
```{r, fig.height=3.5}
ggplot(survey, aes(x=Question.12,y=Question.15)) + 
  geom_point(aes(shape=Question.16), size=3) 
```

## Scatterplot {.build}

Change symbol color by `major` (Q16): requires `aes` argument!


```{r, fig.height=3.5}
ggplot(survey, aes(x=Question.12,y=Question.15)) + 
  geom_point(aes(color=Question.16), size=3) 
```

## Scatterplot {.build}

Change symbol size by number of tv hours/week (Q14): requires `aes` argument!
```{r, fig.height=3.5}
ggplot(survey, aes(x=Question.12,y=Question.15)) + 
  geom_point(aes(color=Question.16, size=Question.14)) 
```

## Scatterplot (legend labels) {.build}
- Need to address labels by type of variable
    - The `color` is the `discrete` variable "major"
    - The `size` is the `continuous` variable tv
```{r, fig.height=3.0}
ggplot(survey, aes(x=Question.12,y=Question.15)) + 
  geom_point(aes(color=Question.16, size=Question.14)) + scale_color_discrete(name="major") + 
  scale_size_continuous(name="tv hours/week")
```

