---
title: "`ggplot` bar graphs, maps, and colors" 
author: "ECON 122"
date: "Day 5"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    keep_md: false
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, warning=FALSE, error=FALSE)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","mapproj","dplyr","scales")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Bar graphs {.build}
```{r, echo=FALSE}
survey <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/Survey.csv")
```
- So far our main variable of interest has been continuous (GPA)
- How do we plot data that is categorical?
- We want to analyze political comfort level (Q9) by religious group (Q8)?
```{r, fig.height=3.5, fig.width=8}
survey <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/Survey.csv")
head(survey$Question.9)
head(survey$Question.8)
```


## Bar graphs {.build}

- We can use `geom_bar`
```{r, fig.height=3.5, fig.width=8}
ggplot(survey, aes(x=Question.8, fill=Question.9)) + geom_bar() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

## Bar graphs {.build}

- The default for `geom_bar` is to sum all observations
- If we want proportions instead use `position="fill"`
```{r, fig.height=3.5, fig.width=8}
ggplot(survey, aes(x=Question.8, fill=Question.9)) + geom_bar(position="fill") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```


## `geom_bar` from summarized data {.build}

- The last example plotted survey response data using `geom_bar`
    - each row represented `one` response
    - `ggplot` automatically summarized them for us
- What if our data is already summarized? 
    - Below is school district passing percentages by year:
```{r, echo=FALSE, fig.height=4.5, fig.width=5}
mca <- data.frame(percent = c(24.5, 33.7, 19.4, 22.4, 27.6, 32.1, 20.2, 20.2, 26.0, 33.7, 20.1, 20.1), meets = factor(rep(c("Exceeds", "Meets", "Partially Meets", "Does not Meet"), 3), levels=c( "Exceeds","Meets","Partially Meets","Does not Meet")), year = rep(c("2015","2016","2017"), each=4))
mca %>% as.tbl
```

## `geom_bar` from summarized data {.build}

- the `percent` variable tells us bar heights (`y`)
- must be paired with `stat="identity"` in `geom_bar`
    - alternatively, can use `geom_col`
```{r, fig.height=3.1}
ggplot(mca, aes(x=year, y=percent, fill=meets)) +
  geom_bar(stat = "identity", position = "fill")
```


## `ggplot2` maps {.build}

The `ggplot2` package contains latitude and longitude to define geographic boundries
    - some regions: `state`, `usa`, `world`, `county`
    - see `maps` package for more regions
```{r}
MainStates <- map_data("state")
str(MainStates)
```

## Maps using `geom_path` {.build}

`geom_path` connects the dots between lat (`y`) and long (`x`) points in a given `group`
```{r, fig.height=3.7}
ggplot(MainStates) + 
    geom_path(aes(x=long, y=lat, group=group))
```

## Maps using `geom_polygon` {.build}

`geom_polygon` is like `geom_path` but it connects start and end points which allows you to `fill` a closed polygon shape
```{r, fig.height=3.7}
ggplot(MainStates) + 
    geom_polygon(aes(x=long, y=lat, group=group), color="black", fill="lightgreen")
```

## Spatial data {.build}

- Earth is a 3-D object but a map is 2-D
    - spatial projection transforms (projects) lat/long locations on a sphere to a cartesian plane
    - distance between latitude is constant (~69 miles/degree lat)
    - distance between longitude decreases as you approach the poles (~69 miles/degree at equator to 0 miles at poles)
- many possible projections that preserve some feature
    - area, angle, distance, ...
- Mercator projection preserves angle but distorts area 

## Maps using `coord_quickmap()` {.build}

the `coord_quickmap` function provides a Mercator approximation by fixing a lat/long ratio
    - good for small regions close to equator
```{r, fig.height=3.7}
ggplot(MainStates) + 
    geom_polygon(aes(x=long, y=lat, group=group), color="black", fill="lightgreen") + 
  coord_quickmap()
```

## Maps using `coord_map()` {.build}

the `coord_map` function provides a Mercator projection
    - see `mapproj` package for more projection options
```{r, fig.height=3.7}
ggplot(MainStates) + 
    geom_polygon(aes(x=long, y=lat, group=group), color="black", fill="lightgreen") + 
  coord_map()
```

## World with `coord_quickmap` {.build}
```{r, fig.height=3.7}
world <- map_data("world")
ggplot(world) + 
    geom_polygon(aes(x=long, y=lat, group=group), color="black", fill="white") + 
  coord_quickmap()
```

## World with `coord_map` {.build}
```{r, fig.height=3.7}
ggplot(world) + 
    geom_polygon(aes(x=long, y=lat, group=group), color="black", fill="white") + 
  coord_map() + scale_x_continuous(limits=c(-180,180))
```

## Cloropleth maps using `geom_map` {.build}

- A cloropleth map uses color or shading of subregions to visualize data
- State level data from the 2016 American Community Survey 
```{r}
ACS <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/ACS2016.csv")

ACS <- ACS[ACS$region != "Alaska" & ACS$region != "Hawaii",] # only 48+D.C.
ACS$region <- tolower(ACS$region)  # lower case (match MainStates regions)
str(ACS)
```

## Cloropleth maps using `geom_map` {.build}

- Want to visualize "% born in state"
- Don't need to merge data. Can have two separate data files: state data `ACS` + mapping data `MainStates`
    - But need to make sure the state names in `region` are the **same**
```{r, fig.height=2.5}
ggplot() + coord_map() + 
  geom_map(data=ACS, aes(map_id = region, fill = PercentInState), map = MainStates) + expand_limits(x=MainStates$long, y=MainStates$lat) + ggtitle("% Born in State")
```

## Cloropleth maps using `geom_map` {.build}

- Add point layer for state capitals in third data set 
```{r, fig.height=3.0}
capitals <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/capitals.csv")
capitals <- capitals[capitals$state != "Alaska" & capitals$state != "Hawaii",] # only 48
ggplot(data=ACS) + coord_map() +
  geom_map(data=ACS, map=MainStates, aes(map_id = region, fill = PercentInState)) + 
   geom_point(data=capitals, aes(x=long,y=lat), size=.7, color="red") + 
  expand_limits(x=MainStates$long, y=MainStates$lat) + ggtitle("% Born in State")
```

## Adjusting color {.build}

- When `color=` is a categorical variable, can use `scale_color_brewer()` or `scale_color_manual()` to change color 
    - same for `fill` colors (change `color` to `fill`)
    - See textbook page 21 for Brewer color palette names
    - can use base-R colors in `rainbow`, `colors` or `palette` for `manual`
- When `color` is a numeric variable, can use `scale_color_distiller()` for Brewer colors and versions of `scale_color_gradient()` for manual colors
    - sequential colors: for low to high data (one direction)
    - diverging colors: for data with a "middle" (higher or lower than the middle)

## Color for numeric variables {.build}

```{r, fig.height=3.7}
mymap <- ggplot() + coord_map() + 
  geom_map(data=ACS, aes(map_id = region, fill = PercentInState), map = MainStates) +
  expand_limits(x=MainStates$long, y=MainStates$lat) + ggtitle("% Born in State") 
mymap + scale_fill_distiller()
```

## Color for numeric variables {.build}

- See `?scale_fill_distiller` for `palette` option names
```{r, fig.height=3.7}
mymap + scale_fill_distiller(palette = "Oranges")
```

## Color for numeric variables {.build}

- Manual choice, see `colors()` for color options
```{r, fig.height=3.7}
mymap + scale_fill_gradient(low="lightpink", high="purple4")
```

## Color for numeric variables {.build}

- Sometimes we want to look at how numbers deviate from 0
- Let's look at the difference between state median income and national median income
    - national median income is estimated to be about $27,000 in 2016
```{r, fig.height=3.7}
mymap2 <- ggplot() + coord_map() + 
  geom_map(data=ACS, aes(map_id = region, fill = MedianIncome - 27000), map = MainStates) +
  expand_limits(x=MainStates$long, y=MainStates$lat) + ggtitle("Difference from US median income") 
```

## Color for numeric variables {.build}

- Sequential choice (default) doesn't highlight deviations from overall median income
```{r, fig.height=3.7}
mymap2 + scale_fill_distiller()
```

## Color for numeric variables {.build}

- `gradient2` manually fills diverging colors, can fix mid color
```{r, fig.height=3.7}
mymap2 + scale_fill_gradient2(midpoint = 0, low="red", high="blue", name="Difference ($)")
```

