---
title: "Data wrangling: multiple data tables" 
author: "ECON 122"
date: "Day 7"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    fig_height: 4
    fig_width: 7
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL, error = TRUE)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```


## Merging data frames {.build}

- Many organizations store various types of data in different sources 
    + different "spreadsheets" or databases
    + think about all the info the college has about you!
        + registrar (grades, courses); Dean of Student (personal info); Admissions (HS, application info)
- We need a common **key** that can ID the same individual across multiple data tables/bases. 
    + name, ID number are common
    + anything that uniquely IDs your units of analysis
- Today: joining/merging tables using a common key


## Different types of joins {.build}

-  Mutating joins - combine vars from both tables
    - `inner_join(x,y, by=c("x_key"="y_key"))`: returns all rows in `x` that have a match in `y` 
    - `left_join(x,y, by=c("x_key"="y_key"))`: returns all rows in `x`, will fill `NA` in `y` columns with no match
    - `right_join(x,y, by=c("x_key"="y_key"))`: returns all rows in `y`, will fill `NA` in `x` columns with no match
    - `full_join(x,y, by=c("x_key"="y_key"))`: returns all rows from both `x` and `y`, will fill `NA` in columns with no match

- Filtering joins - keep variables from left-hand table
    - `semi_join(x,y, by=c("x_key"="y_key"))`: returns all rows from `x` where there are matching values in `y`, keeping just columns from `x`
    - `anti_join(x,y, by=c("x_key"="y_key"))`: returns all rows from `x` where there are **not** matching values in `y`, keeping just columns from `x`

## Example: class grades {.build}

- I want to know how people are performing in this class relative to their average GPA
- Load dataframe with class grades
```{r}
class_grades <- data.frame(student_ID=1:5,
                           grade=c("A","A+","B-","B+","A-"),
                           college=c("PIT","CMC","CMC","CMC","SRC"))
class_grades 
```

## Example: class grades {.build}

- I can get access to CMC students' GPAs but not the other colleges

```{r}
GPA <- data.frame(CMC_ID=c(2:4,10:11),gpa=c(2.7,4.0,3.1,3.6,3.7))
GPA 
```

## Example: class grades {.build}

- How should I join the two tables?
    - `inner_join` seems like a good solution
- What keys to use? 
    - The keys have different variable names but that's OK

```{r}
inner_join(class_grades,GPA,by = c("student_ID"="CMC_ID")) 
```
- What are the downsides to this inner join?  
    - I don't see my PIT and SRC students anymore
- What if I want to compare grades **within** the class as well as **GPAs**?
    
## Example: class grades {.build}

- Let's try a `left_join` to keep all students
```{r}
left_join(class_grades,GPA,by = c("student_ID"="CMC_ID")) 
```
- Left join will never **remove** information from the left dataframe
    - Perhaps the fact that there **isn't** a match is important
- Because I can't access PIT and SRC data, GPAs will be `NA` for those students

## Example: class grades {.build}

- What will a `right_join` give us?
```{r}
right_join(class_grades,GPA,by = c("student_ID"="CMC_ID")) 
```
- All the students that I have GPA information for, even if they are not in my class
- Note: `right_join` is just `left_join` but with the order switched
- Don't *really* need it, but it can be useful if you don't want to switch order
  
  
## Example: class grades {.build}

- What does `full_join` do?
```{r}
full_join(class_grades,GPA,by = c("student_ID"="CMC_ID")) 
```
- All possible information
- Dangerous if you have two very large tables


## Example: class grades {.build}
- What if I just want to know which rows have a match?
    + Ex: my college variable got deleted
```{r}
class_grades %>% select(-college) 
```
```{r}
class_grades %>% select(-college) %>% semi_join(GPA,by = c("student_ID"="CMC_ID"))
```
- Only keeps variables from the left table (no `GPA`)
- Only returns rows which have a match
    - Similar to `inner_join` but will never duplicate rows of left table

## Example: class grades {.build}

- What if I just want to know which rows **don't** match
```{r}
anti_join(class_grades,GPA,by = c("student_ID"="CMC_ID")) 
```
- Tells me which students don't have a GPA match (BC they are not CMC students)
- Sometimes identifying which records don't match is the goal itself 
    - For example, may tell you which years, states, students are missing so you can request that data
    
## More examples {.build}

- So far: One unique ID where each ID only shows up once in the table
- Other scenarios
    - What if we have a table that has multiple observations per unique ID? 
        + multiple classes per student
    - What happens if we want to merge on multiple variables? 
        + merge on student ID and semester
- The activity today introduces the above two scenarios
    