---
title: "Data intake and Dates" 
author: "ECON 122"
date: "Day 8"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    fig_height: 4
    fig_width: 7
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

```{r packageCheck, include=FALSE}
mypacks <- c("dplyr","readr","lubridate")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
local_edition(1)
```

## Data intake

- Data from the real world may not always come in the form you expect
- The process of importing data itself may already lead to false conclusions
    - Data intake in R makes assumptions about how to parse raw data
- We should always view the raw data before importing
    - For smaller files, spreadsheets are useful
    - For larger files, `unix` commands like `head` or `more` are useful 

## Factor vs String: {.build}

- Recall: factors (coded internally as integer levels) take less storage than strings
- Sometimes mostly numeric variables contain a few character entries. 
- What happens when we coerce into a numeric variable?

## String into numeric: {.build}

- Character string variable
```{r}
(x <- c("4", "9","2.1","oops") )
typeof(x)
```
- Coerce into numeric with `as.numeric`
```{r}
(y <- as.numeric(x))
typeof(y)
```
- Coercion correctly converts to numeric vector with NA for non-numeric

## Factor into numeric: {.build}

- Factor  variable
```{r}
(x <- factor(c("4", "9","2.1","oops")) )
typeof(x)
```
- Coerce into numeric with `as.numeric`
```{r}
(y <- as.numeric(x))
typeof(y)
```
- Coercion is incorrect: returns integer level value, not variable values

## Factor into numeric: {.build}

- Can get correct coercion with `readr::parse_number`
```{r}
y <- parse_number(as.character(x))  # vector with "problem" attribute
typeof(y)
as.vector(y)  # plain vector
```
- Coercion is now correct

## `read_csv` vs. `read.csv`:  {.build}

- the base R `read.csv` default setting makes all columns with characters a factor:
    + default: `stringsAsFactors=TRUE`
    +  Fixed in the most recent version but can still cause issues with older versions
```{r e1}
energy1 <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv",stringsAsFactors=TRUE)
glimpse(energy1[,c("Timestamp","Musser_Hall")])
```
- BAD: Default reads `Timestamp` is a factor 
    + levels will be ordered numerically/alphabetically not chronologically
- GOOD: Musser Hall energy values are `dbl` (double)
  
## `read_csv` vs. `read.csv`:  {.build}

- the `readr` package `read_csv` default setting makes all columns with characters a character:
- Default: it uses the first 1000 rows to guess variable types (`guess_max`)
    - Fixed in the most recent version but can still cause issues with older versions
```{r e2, warning=FALSE}
local_edition(1)
energy2 <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv")
glimpse(energy2[,c("Timestamp","Musser_Hall")])
```
- GOOD: Default reads `Timestamp` is a date/time type called `POSIXct` 
- BAD: Musser Hall energy values are `logical` 
  
## `read_csv` vs. `read.csv`:  {.build}

- Summaries for `energy1` and `energy2` are not the same!
```{r}
summary(energy1$Musser_Hall)
summary(energy2$Musser_Hall)
```
- The first 22,243 rows in `energy1` are NA before they become numerical
```{r}
energy1[22238:22245,"Musser_Hall"]
```


## `read_csv` vs. `read.csv`:  {.build}

- If `read_csv` guesses column type wrong you get a warning message:
```{r}
problems(energy2)
```

## `read_csv` vs. `read.csv`:  {.build}

- Solution: 
    + make sure you have the latest version
    + check for warnings, define column types if needed
    + or increase `guess_max` value

## `read_csv` vs. `read.csv`:  {.build}

- We can also explicitly specify what storage type we want
- `Timestamp` is a `datetime` variable
- `dayWeek` is a factor with ordered levels
- all other read in as `double`
```{r}
energy <- read_csv("https://raw.githubusercontent.com/mgelman/data/master/EnergyData1516.csv", 
                   col_type = cols(.default = col_double(), 
                          Timestamp = col_datetime(format = ""),
                          dayWeek = col_factor(
                              levels=c("Mon","Tues","Wed","Thurs","Fri","Sat","Sun"))))
```
- No more warning messages after reading

## Dates and Times:  {.build}

- R has  built-in functions to handle dates and times
- Stores dates, times, or date-times as either:
    + `POSIXct` type: time in seconds since start of 1970
    + `POSIXlt` type: list of vector of time/date units
- The `lubridate` package makes dates and times manipulations a little easier

## `lubridate`: extract info {.build}

- Let's look at the 5th `Timestamp` in the energy data:
```{r}
( stamp5 <- energy$Timestamp[5] )
str(stamp5)
```
- What day of the week and day of the year was it?
```{r}
wday(stamp5, label=TRUE)
yday(stamp5)
```

## `lubridate`: extract info {.build}

- We can also extract date and time info:
```{r}
stamp5
year(stamp5)
month(stamp5)
day(stamp5)
hour(stamp5)
minute(stamp5)
```

## `lubridate`: create dates and times {.build}

- There are many variations of commands like:
```{r}
(mydate <- mdy("1/24/2018") )
ymd("2018-01-24")
ymd_hm("2018-01-24 13:32")
```
- date part: y=year, m=month, d=day
- time part: h=hour, m=minute, s=second

## `lubridate`: create dates and times {.build}

- Also `make_date` or `make_datetime`
```{r}
energy %>% mutate(datetime = make_datetime(year,month,dayOfMonth,timeHour,timeMinute)) %>%
  select(Timestamp, datetime, year,month,dayOfMonth,timeHour,timeMinute)
```

## `lubridate`: updating dates and times {.build}

- Can modify an existing date with `update` 
```{r}
mydate
update(mydate, year = 2020)
update(mydate, hour=12, minute = 23)
```

## `lubridate`: measuring time {.build}

- Suppose we want to see how much time has elapsed since the start of the term:
```{r}
firstDay<- mdy("08/28/2023")
firstDay
```
- We can get a date/time with `now()` and a date with `today()`
```{r}
today()
now()
```

## `lubridate`: measuring time {.build}

- `interval` creates an "interval" class object that defines an interval of time (tied to specific dates)
```{r}
interval(firstDay, today())  
```
- `duration` specifies a length of time (not tied to specific dates)
```{r}
dminutes(2)
dhours(1)
ddays(1)
```

## `lubridate`: measuring time {.build}

- We use an interval and duration to compute a length of time
- How many days so far this term?
```{r}
interval(firstDay, today()) / ddays(1) 
```
- How many weeks? minutes?
```{r}
interval(firstDay, today()) / dweeks(1)
interval(firstDay, today()) / dminutes(1)
```

## `lubridate`: measuring time {.build}

- Can also use date/time variables:
- Difference between first two time stamps:
```{r}
energy$Timestamp[1:2]
interval(energy$Timestamp[1], energy$Timestamp[2])/dminutes(1)
```
- 15 minutes between readings 1 and 2

