---
title: "Tidy data: restructuring" 
author: "ECON 122"
date: "Day 9"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    fig_height: 4
    fig_width: 7
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","babynames","dplyr","tidyr","readr")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Two main ways to structure a dataset  {.build}

1. `wide` data typically represent unique variables via columns (in extreme cases, may have more columns than rows) 
2.  `narrow\long` data represent values in columns but each row represents a different case (typically has more rows than columns)
- Today we will learn how to convert a dataset from one type to the other and vice versa 

## Typical "spreadsheet" data: wide `babynames`  {.build}

```{r,echo=FALSE}
babyname_wide <- babynames%>% filter(name %in% c("Anna","Mary","James")) %>% select(-prop) %>% spread(key=year,value=n) 
babyname_wide[1:5,1:10]
```
- rows: unique name/sex combo
- columns: many, many columns containing the same type of value measured in a different year

- When is this format useful? 


##  tidy data: narrow `babynames` {.build}

<div class="columns-2">
```{r,echo=FALSE}
babynames %>% print(n=8)
```


- rows: each `case` refers to unique name/sex/year combo
- columns: `variables` containing the same type of value for each row
    + count/prop for each case
- when is this format useful? 
</div>


## Tidy Data 

- more long than wide
- Pros:
    + scalable, easier to add more variables and join other tables
        - How would `prop` be added to "wide" `babynames` format?
    +  more efficient storage/retrieval 
    + usually easier to analyze (graph, model, etc)
    
- Cons: 
    + harder to "see" all data in spreadsheet
- Management of data
    + done via code, not within spreadsheet app
    + fits our philosophy of "reproducibility"  
    

## Reshaping data: wide format {.build}

- Here are 2014 and 2015 grad rates for 3 schools (cases):

```{r, echo=FALSE}
grad_wide<- data.frame(school=c("Aitkin","Bemidji","Chaska"), gradRate_2015=c(93.0,86.2,96.2),gradRate_2014=c(94.6,83.2,92.7),region=c("northern","northern","metro"))
grad_wide
```

- We want to analyze graduation trends by year

- How would we plot this data?
    - `ggplot(data,aes(x=year,y=gradRate))`

- We can convert data between wide and narrow (long) formats to suit our analysis needs

## Reshaping data: from wide to narrow {.build}

- The `tidyr` package let's us go between narrow and long formats
- `gather(wide, key=, value=, ..cols to gather..)`
    + `key` is the **(new)** variable name that will ID cases
    + `value` is the **(new)** variable name given to the gathered values
    + `cols to gather`: comma separated names of columns to gather
- any variables that are not gathered will have values repeated

## Reshaping data: wide to narrow  {.build}

- Our values to fill are grad rates and our cases will be defined by school/year combinations so `year` is the `key` (`school` is already a variable):
```{r, echo=TRUE}
grad_narrow <- gather(grad_wide, key = year, value=gradRate, gradRate_2015, gradRate_2014)
grad_narrow
```
- The year variable is not quite a number though...

## Reshaping data: `parse_number` {.build}

- We can use the `readr` package's command `parse_number` to clean up `year` levels
    - drops non-numerics before/after first numeric value(s)
```{r}
grad_narrow <- grad_narrow %>% mutate(year=parse_number(year))
grad_narrow
```

## Plotting grad rates

```{r}
ggplot(grad_narrow, aes(x=year,y=gradRate)) + geom_line(aes(color=school))+
  geom_point(aes(color=school,shape=school)) + labs(x="Year",y="Graduate Rate") +
  scale_x_continuous(breaks=c(2014,2015))
```

## Reshaping data: wide format {.build}

- What about `wide` data with two time varying variables? 
- Here are 2014 and 2015 4- and 6-year grad rates for 3 schools (cases):

```{r, echo=FALSE}
grad_wider<- data.frame(school=c("Aitkin","Bemidji","Chaska"), gradRate4_2015=c(93.0,86.2,96.2),gradRate4_2014=c(94.6,83.2,92.7),region=c("northern","northern","metro"), gradRate6_2015 = c(93.5,90.3, 98.4), gradRate6_2014 = c(94.5, 88.4, 97.8))
grad_wider %>% as_tibble()
```


## Reshaping data: wide to narrow  {.build}

- Our values to fill are grad rates and our cases will be defined by school/year combinations so `year` is the `key`:
```{r, echo=TRUE}
grad_narrower <- gather(grad_wider, key = year, value=gradRate, c(2,3,5,6))
grad_narrower %>% as_tibble() %>% print(n=7)
```
- What is the next step before calling `ggplot`? 

## Reshaping data: wide to narrow  {.build}

- There are *two* variables contained in `year`: 4- vs. 6- year rate and year
    - `tidyr` package `separate` command
```{r}
grad_narrower <- grad_narrower %>% separate(year, into=c("type","year"), sep="_")
grad_narrower %>% as_tibble()
```

## Plotting grad rates (6 year) {.build}

```{r}
grad_narrower %>% filter(type=="gradRate6") %>%
ggplot(aes(x=as.numeric(year),y=gradRate)) + geom_line(aes(color=school))+
  geom_point(aes(color=school,shape=school)) + labs(x="Year",y="6 year Graduation Rate") +
  scale_x_continuous(breaks=c(2014,2015)) 
```

## Plotting grad rates (layer using `linetype`){.build}

```{r, fig.height=3.8,fig.width=7}
ggplot(grad_narrower, aes(x=as.numeric(year),y=gradRate)) + 
  geom_line(aes(color=school, linetype=type))+
  geom_point(aes(color=school,shape=school)) + labs(x="Year",y="Graduate Rate") +
  scale_x_continuous(breaks=c(2014,2015)) + 
  scale_linetype_discrete(labels=c("4 year","6 year"))
```

## Reshaping data: narrow to wide {.build}

- the `spread` command can convert narrow data into wide data
- `spread(narrow, key=, value=)`
    + `key` is the (existing) variable that will be used to name the new variables
    + `value` is the (existing) variable whose values will fill the new columns
    + if `key` has numeric values, add `sep="_"` to get the `key` row name attached to numeric values
```{r}
grad_narrow
```
## Reshaping data: narrow to wide {.build}

```{r}
spread(grad_narrow, key=year, value=gradRate)
```
- columns with numeric names are more difficult to work with so add a `sep` argument:
```{r}
spread(grad_narrow, key=year, value=gradRate, sep="_")
```

## Reshaping data: narrow to wide {.build}
- What if two variables (`type` and `year`) identify grad rates

```{r}
grad_narrower
```


## Reshaping data: narrow to wide {.build}

- one way: `unite` into one variable
- default separator is `"_"`

```{r}
grad_narrower2 <-  unite(grad_narrower, key, type, year) 
grad_narrower2
```

## Reshaping data: narrow to wide {.build}

- Then spread with the united variable as key
```{r}
spread(grad_narrower2, key=key, value=gradRate) %>% as_tibble()
```
